---
title: "STUDY — power / sample size for the primary endpoint"
subtitle: "One line on the design"
author: "Your name"
date: "Month Year"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: kable
---

<!--
Copy this directory to analyses/YYYY-MM_study-slug/ and work through it.
The headings are the ones a reviewer, a DSMB, or whoever inherits this in
two years will look for, in roughly that order. Delete what doesn't apply,
but think twice before deleting "Verify the DGP" or "Validation".

When you're done:
  source("../../R/load.R")
  add_entry(id = "YYYY-MM_study-slug", title = ..., study = ..., ...)
  build_index()
-->

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 4.5)
source("../../R/load.R")
library(knitr); library(ggplot2)
```

## The question

State, in plain sentences a clinician will read:

- **Design**: how many arms, how randomised, what unit.
- **Primary endpoint**: exactly as it will be defined in the protocol.
- **Effect worth detecting**, and where that number came from (pilot data,
  prior trial, clinical judgement — say which).
- **Budget or constraint**: fixed N? fixed number of sites? fixed follow-up?
- **Decision rule**: what counts as success. `p < 0.05`? A posterior
  probability bar? A non-inferiority margin?
- **Multiplicity**: how many comparisons are confirmatory, and what the
  adjustment is.

If a number here is a guess, say it is a guess. That is the assumption the
sensitivity analysis below exists to stress.

## Design

```{r design}
d <- design_binary_parallel()   # or whichever design_*() fits; see ../../README.md
d
```

If no existing design fits, add a `design_*()` constructor to `R/` rather
than defining the DGP inline here. Inline DGPs are how we ended up with a
folder of unshareable scripts.

```{r params}
pars <- design_params(d,
  # the assumed truth
)
```

## Verify the DGP

Confirm the simulator produces the effect it was asked for, *before*
powering anything. One huge simulated trial, and recover the parameter:

```{r dgp-check, cache = TRUE}
set.seed(1)
# big <- d$dgp(design_params(d, n = 200000))
# ... estimate the effect from `big` and compare it to the assumed truth
```

This catches parameterisation slips — the failure mode where the simulation
runs fine and cleanly powers the wrong effect size.

## Primary power result

```{r power, cache = TRUE}
# r <- run_power(d, grid, n_sim = 2000, seed = 2026)
```

```{r power-table, echo = FALSE}
# kable(...)  <- presentation goes in its OWN chunk
```

Two conventions worth keeping. Report `n_failed` and the Monte Carlo SE
alongside the power, so it's clear how much of the last digit is real. And
keep the `kable()` out of the cached chunk: if the table lives inside it,
every wording change re-runs the simulation, which for a JAGS design means
waiting twenty minutes to fix a typo.

## Validation

```{r validate, cache = TRUE}
# check_against_analytic(d, grid, n_sim = 4000)
```

If the design has a closed form, check against it and show the comparison.
If it doesn't, say so explicitly and name what you checked instead — a
degenerate case, a limiting case, agreement with published numbers, or the
type I error under the null:

```{r null, cache = TRUE}
# Type I error: same machinery, with the effect set to zero.
# run_power(d, <grid with no effect>, n_sim = 8000)
```

## Sensitivity

At minimum, vary the thing you were least sure about in "The question".
Common candidates: the control-arm rate, the ICC, the effect size, the
prior, the dropout rate, the accrual assumption.

```{r sens, cache = TRUE}
# rs <- run_power(d, expand.grid(...), n_sim = 1000, seed = 3026)
```

```{r sens-plot, echo = FALSE}
# ggplot(rs, aes(x, power, colour = factor(z))) +
#   geom_hline(yintercept = 0.80, linetype = 2, colour = "grey40") +
#   geom_ribbon(aes(ymin = power_lo, ymax = power_hi, fill = factor(z)),
#               alpha = 0.15, colour = NA) +
#   geom_line() + geom_point() +
#   scale_y_continuous(limits = c(0, 1)) +
#   theme_minimal(base_size = 11)
```

## Headline

The sentence that goes into the application, with every number filled in
from the objects above rather than typed by hand — so it can't drift out of
sync when the assumptions change.

```{r headline, results = "asis", echo = FALSE}
# cat(sprintf("> With N = %d ..., the trial has %.0f%% power to detect ...", n, 100 * r$power))
```

## Assumptions and limitations

List them. The ones that are always worth a line:

- Where the control-arm parameter came from.
- Dropout / loss to follow-up, and whether it is in the simulation or
  handled by inflating N afterwards.
- Whether the analysis model in the simulation matches the one in the SAP.
- Anything the simulation does *not* represent (site effects, interim
  analyses, missing data, non-adherence).

## Reproducibility

```{r session, echo = FALSE}
sessionInfo()$R.version$version.string
```

Seeds are set in every `run_power()` call above, so re-knitting reproduces
these numbers exactly.
