---
title: "PARMA — Bayesian power for a Weibull time-to-event endpoint"
subtitle: "Time to resolution of hypoxemia, high vs. low driving pressure in pediatric ARDS"
author: "Zhe Chen, with Nadir Yehya"
date: "September 2025 (ported into the power library, September 2026)"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: kable
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 4.5)
source("../../R/load.R")
library(knitr); library(ggplot2)
```

## The question

PARMA randomises **160** children with ARDS to a high vs. low driving pressure
ventilation strategy. The primary endpoint is **time to resolution of
hypoxemia** (alive with PaO₂/FiO₂ > 300 on two measurements four hours apart),
followed for **28 days**. Death acts as a **competing event**.

Because the event here is a *good* one, a hazard ratio **greater than 1** is
benefit: the treated arm resolves sooner.

Inference is Bayesian, and the design question is correspondingly not "will
p < 0.05" but:

> How often would this trial produce a posterior probability of benefit,
> `Pr(HR > 1 | data)`, above 0.90?

That frequency **is** the power. In this library it is the same
`run_power()` call as any frequentist design — the only thing that changes is
what the design's `analyze()` returns as `reject`.

```{r design}
d <- design_weibull_bayes()
d
```

## Data-generating model

Event times are Weibull. The control arm is calibrated to empirical CHOP data;
the treated arm shifts it by the target log hazard ratio.

```{r dgp-params}
pars <- design_params(d,
  n              = 160,
  beta           = log(1.6),   # target log HR (benefit)
  shape_event    = 0.9,
  scale_event    = 4,
  competing_dist = "exp",
  scale_comp     = 21,         # mean time to the competing event (death)
  horizon        = 28,
  threshold      = 0.90
)

ctrl <- weibull_control_summary(pars$scale_event, pars$shape_event)
trt_scale <- weibull_scale_for_arm(pars$scale_event, pars$shape_event, pars$beta, 1)
trt  <- weibull_control_summary(trt_scale, pars$shape_event)

kable(data.frame(
  Arm    = c("Control", "High driving pressure"),
  `Weibull scale` = sprintf("%.2f", c(pars$scale_event, trt_scale)),
  `Mean time to resolution (days)`   = sprintf("%.1f", c(ctrl["mean"],   trt["mean"])),
  `Median time to resolution (days)` = sprintf("%.1f", c(ctrl["median"], trt["median"])),
  check.names = FALSE
), caption = "Implied control and treated arms. Shape = 0.9, true HR = 1.6.")
```

### The parameterisation, written down once

This is the step that quietly breaks these simulations, so it lives in one
function and is stated explicitly here.

R's `rweibull(shape = a, scale = b)` has $S(t) = \exp\{-(t/b)^a\}$.
JAGS's `dweib(alpha, lambda)` has $S(t) = \exp\{-\lambda t^{\alpha}\}$.
They match with $\alpha = a$ and $\lambda = b^{-a}$. Proportional hazards with
log-HR $\beta$ therefore requires

$$b_{\text{trt}} = b_{\text{ctrl}} \cdot e^{-\beta / a},$$

which is `weibull_scale_for_arm()`. The consequence of the exponent is that the
baseline intercept on the JAGS scale is $\beta_0 = -a \log b_{\text{ctrl}}$:

```{r beta0}
sprintf("beta0 implied by the data-generating control arm: %.3f",
        weibull_beta0_true(pars$scale_event, pars$shape_event))
```

That matters below, because the baseline prior is centred on it.

### Verifying the DGP hits the intended effect

Before powering anything, confirm the simulator produces the hazard ratio it
was asked for. Fit a Cox model to one very large simulated trial:

```{r dgp-check, cache = TRUE}
set.seed(1)
big <- d$dgp(design_params(d, n = 200000))
cf  <- summary(survival::coxph(survival::Surv(time, status) ~ trt, data = big))$coefficients
sprintf("target log HR = %.4f | Cox estimate on n = 200,000: %.4f (SE %.4f) | event fraction %.3f",
        log(1.6), cf["trt", "coef"], cf["trt", "se(coef)"], mean(big$status))
```

```{r dgp-plot, echo = FALSE, fig.height = 4}
set.seed(2)
one <- d$dgp(pars)
km  <- survival::survfit(survival::Surv(time, status) ~ trt, data = one)
plot(km, col = c("#4C6EF5", "#E8590C"), lwd = 2, xlab = "Days since randomisation",
     ylab = "Proportion with unresolved hypoxemia", mark.time = TRUE)
legend("topright", c("Low driving pressure (control)", "High driving pressure"),
       col = c("#4C6EF5", "#E8590C"), lwd = 2, bty = "n")
title("One simulated PARMA trial (n = 160)")
```

## Analysis model

```{r jags-model, echo = FALSE, comment = ""}
cat(readLines(powerlib_path("jags", "weibull_ph.jags")), sep = "\n")
```

Note that `N` and every prior hyperparameter arrive as data. The original
version of this model hard-coded `for(i in 1:160)`, which meant a new file per
sample size; making them data is what lets the same file serve the sensitivity
analyses below.

Priors, as in the original analysis plan:

| Parameter | Prior | Rationale |
|---|---|---|
| `beta.trt` | `dnorm(0, 0.2)` | Minimally informative, centred at **no effect** (HR = 1). Precision 0.2 = SD ≈ 2.24 on the log-HR scale. |
| `beta0` | `dnorm(·, 0.5)` | Baseline hazard, moderately informative, centred on the control-arm value implied by the empirical data. |
| `alpha` | `dgamma(1.1, 1.1)` | Weibull shape; mean 1, variance ≈ 0.91. |

Centring `beta.trt` at zero is what makes the power calculation honest: the
prior is not helping the trial find an effect.

## Power

```{r power-main, cache = TRUE}
r <- run_power(d, data.frame(n = 160), n_sim = 1000, seed = 2026, verbose = FALSE)

kable(data.frame(
  Quantity = c("Power: Pr(HR > 1) >= 0.90",
               "Pr(HR > 1) >= 0.90 against the *assumed true* HR of 1.6",
               "95% credible interval excludes HR = 1",
               "Mean posterior Pr(HR > 1)",
               "Mean posterior log HR (truth = 0.470)",
               "Mean event fraction by day 28"),
  Value = sprintf("%.3f", c(r$power, r$reject_vs_truth, r$reject_ci95,
                            r$prob_benefit, r$post_mean_beta, r$event_frac)),
  `MC SE` = c(sprintf("%.3f", r$mc_se), rep("", 5)),
  check.names = FALSE
), caption = "1000 simulated trials, n = 160, true HR = 1.6.")
```

The second row is what the original memo reported as a "type I error rate". It
is the same decision rule applied against the *assumed true* effect rather than
the null — how often the trial would claim more benefit than was actually built
in. It is a useful calibration diagnostic, but it is **not** a frequentist type
I error rate, and it should be described as what it is when it appears in a
protocol.

### Reproducing the original memo

The original PARMA submission quoted 93% power at a type I error of 0.101,
using a baseline prior fixed at `dnorm(-1.6, 0.5)`. The ported pipeline
reproduces it:

```{r repro, cache = TRUE}
rep <- run_power(d, data.frame(
    label = c("This port (beta0 centred on the DGP)",
              "Original memo (beta0 ~ dnorm(-1.6, 0.5))"),
    center_beta0_at_truth = c(TRUE, FALSE),
    prior_beta0_mean      = c(NA, -1.6)),
  n_sim = 1000, seed = 2026, verbose = FALSE)

kable(data.frame(
  Configuration = rep$label,
  Power = sprintf("%.3f (SE %.3f)", rep$power, rep$mc_se),
  `"Type I error"` = sprintf("%.3f", rep$reject_vs_truth),
  check.names = FALSE
), caption = "Memo quoted 93% power and type I error 0.101.")
```

## Sensitivity: sample size and effect size

```{r grid, cache = TRUE}
g <- expand.grid(n = c(120, 140, 160, 180, 200, 240),
                 beta = log(c(1.4, 1.6, 1.8)))
rg <- run_power(d, g, n_sim = 500, seed = 4026, verbose = FALSE)
rg$HR <- exp(rg$beta)
```

```{r grid-plot, echo = FALSE}
ggplot(rg, aes(n, power, colour = factor(round(HR, 1)))) +
  geom_hline(yintercept = c(0.80, 0.90), linetype = 2, colour = "grey50") +
  geom_ribbon(aes(ymin = power_lo, ymax = power_hi, fill = factor(round(HR, 1))),
              alpha = 0.15, colour = NA) +
  geom_line(linewidth = 0.8) + geom_point() +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = "Total sample size", y = "Pr( Pr(HR > 1 | data) >= 0.90 )",
       colour = "True HR", fill = "True HR",
       caption = "500 simulations per point; ribbon = Monte Carlo 95% interval.") +
  theme_minimal(base_size = 11)
```

```{r grid-table, echo = FALSE}
w <- reshape(data.frame(n = rg$n, HR = sprintf("HR = %.1f", rg$HR),
                        power = sprintf("%.2f", rg$power)),
             idvar = "n", timevar = "HR", direction = "wide")
names(w) <- sub("^power\\.", "", names(w))
kable(w, row.names = FALSE, caption = "Bayesian power by sample size and true HR.")
```

## Sensitivity: the prior on the treatment effect

The one assumption a reviewer will press on. Precision 0.2 is the pre-specified
minimally informative choice; the others bracket it.

```{r prior-sens, cache = TRUE}
rp <- run_power(d, data.frame(prior_trt_prec = c(0.05, 0.2, 1, 4)),
                n_sim = 500, seed = 5026, verbose = FALSE)
rp$prior_sd <- 1 / sqrt(rp$prior_trt_prec)

kable(data.frame(
  `Prior precision` = rp$prior_trt_prec,
  `Prior SD (log HR)` = sprintf("%.2f", rp$prior_sd),
  Power = sprintf("%.3f", rp$power),
  `Mean posterior log HR` = sprintf("%.3f", rp$post_mean_beta),
  `Shrinkage vs. truth (0.470)` = sprintf("%+.3f", rp$post_mean_beta - log(1.6)),
  check.names = FALSE
), caption = "Effect of the treatment-effect prior. All centred at HR = 1.")
```

A tighter prior centred on the null shrinks the posterior toward no effect and
costs power — which is the expected direction, and worth showing rather than
asserting.

## Frequentist comparator on identical data

`design_weibull_cox()` shares the same DGP function, so any difference in power
is attributable to the analysis model and not to a differently simulated trial.

```{r cox, cache = TRUE}
rc <- run_power(design_weibull_cox(), data.frame(n = c(120, 160, 200)),
                n_sim = 1000, seed = 6026, verbose = FALSE)
rb <- run_power(d, data.frame(n = c(120, 160, 200)),
                n_sim = 1000, seed = 6026, verbose = FALSE)

kable(data.frame(
  `Total n` = rc$n,
  `Bayesian: Pr(HR>1) >= 0.90` = sprintf("%.3f", rb$power),
  `Bayesian: 95% CrI excludes 1` = sprintf("%.3f", rb$reject_ci95),
  `Cox: p < 0.05 (two-sided)` = sprintf("%.3f", rc$power),
  check.names = FALSE
), caption = "Same simulated trials, three decision rules.")
```

The one-sided 0.90 posterior bar is a materially easier threshold than a
two-sided 5% test — the gap between columns two and four is the design
decision, not a modelling artefact, and it should be presented to the DSMB as
such.

## Notes for reuse

- **Different endpoint direction.** If the event is *bad* (death, readmission),
  benefit is HR < 1: flip the sign of `beta` and read `reject` as
  `Pr(HR < 1) >= threshold` by negating `beta.trt`.
- **Different competing-risk structure.** `competing_dist = "weibull"` with
  `shape_comp`/`scale_comp`, or `"none"` to switch it off.
- **Exponential or piecewise baseline.** The exponential case is
  `alpha` fixed at 1; a piecewise model needs a new `.jags` file but the same
  `analyze()` skeleton.
- **Runtime.** 1000 replicates × one JAGS chain of 5000 iterations is a few
  minutes on 8 cores. Drop `n_iter` while exploring, raise it for the number
  that goes in the application.

```{r session, echo = FALSE}
sessionInfo()$R.version$version.string
```
