---
title: "PARMA Trial: Power Calculation and Bayesian Analysis Plan"
output: 
  pdf_document:
    latex_engine: xelatex
---


## **Overview**

The PARMA trial uses a **Bayesian time-to-event model** to assess the efficacy of a high versus low driving pressure (ΔP) ventilation strategy in pediatric ARDS. The trial's **primary endpoint** is time to resolution of hypoxemia, defined as **alive with PaO₂/FiO₂ > 300** on two measurements 4 hours apart.

### **Key Design Elements**
- **Planned sample size**: 160 patients
- **Primary outcome**: Time to hypoxemia resolution
- **Minimal censoring**: The censoring times are much larger than time-to-event for both groups
- **Follow-up period**: 28 days per participant
- **Model**: Bayesian survival model with Weibull distribution
- **Statistical inference**: Calculate the power to detect a >90% posterior probability of benefit (Hazard Ratio: HR > 1), assuming a true hazard ratio of 1.6. This corresponds to mean times to resolution of approximately 5 days in the control arm and 3 days in the intervention arm.

## **Model Specification**

### **Weibull Bayesian Survival Model (JAGS)**

The primary analysis will employ a Bayesian Weibull survival model implemented in JAGS.  

```jags
model{
  for(i in 1:160) {
    # Interval censoring for competing risks
    is.censored[i] ~ dinterval(t[i], t.cen[i])
    
    # Weibull survival model  
    t[i] ~ dweib(alpha, lambda[i])
    lambda[i] <- exp(beta0 + beta.trt * trt[i])
  }
  
  # Priors
  beta0 ~ dnorm(-1.6, 0.5)      # Centered at 5-day mean for control group
  beta.trt ~ dnorm(0.0, 0.2)    # Uninformative treatment effect (HR = 1)
  alpha ~ dgamma(1.1, 1.1)      # Weibull shape parameter
}
```

#### **Primary Analysis Priors**
- **Treatment effect (`beta.trt`)**: `dnorm(0.0, 0.2)` → minimally informative: mean = 0 (HR = 1), low precision
- **Baseline hazard (`beta0`)**: `dnorm(-1.6, 0.5)` → centered at 5-day mean for control group
- **Shape parameter (`alpha`)**: `dgamma(1.1, 1.1)` → reflects empirical data


## **Power Simulation Strategy**

- Randomize 160 patients to two arms
- Generate event times via Weibull distribution
- Censor at 28 days or competing event
- Estimate posterior distribution of treatment effect (log hazard ratio)

### **Data-Generating Parameters (Based on Empirical Data)**
```r
n <- 160          # Sample size
beta <- log(1.6)  # log HR (projected true effect)
shape1 <- 0.9    # Shape parameter for primary event 
scale1 <- 4   # Scale parameter for primary event
scale2 <- 21  # Scale parameter for competing event
```

```r
# Main simulation function
CHOP_simulation <- function(beta, shape1, scale1, scale2, n) {
  
  # Generate treatment assignment (1:1 randomization)
  tx <- rbinom(n, 1, 0.5)
  
  # Generate primary event times (hypoxemia resolution)
  event <- rweibull(n, shape = shape1, scale = (scale1 * exp(-beta * tx))^(1 / shape1))
  event <- pmin(event, 28)  # Censor at 28 days 
  
  # Generate competing event times (death)
  competing <- rexp(n, rate = 1/scale2)
  competing <- pmin(competing, 28)  # Censor at 28 days
  
  # Determine event status (1 = primary event, 0 = competing/censored)
  Status <- ifelse(event < competing, 1, 0)
  
  # Create simulation dataset
  simdata <- data.frame(
    event = event, 
    competing = competing, 
    Status = Status, 
    tx = tx
  )
  
  # Prepare data for JAGS
  t <- simdata$event
  is.na(t) <- simdata$Status == 0  # Missing for competing events
  
  tinits1 <- simdata$event + 3
  is.na(tinits1) <- simdata$Status == 1
  
  jagsdata <- list(
    t = t,
    t.cen = simdata$competing,
    trt = simdata$tx,
    is.censored = (1 - simdata$Status)
  )
  
  jagsinits <- list(list(t = tinits1, beta0 = -1.6))
  
  # Fit JAGS model
  tryCatch({
    jags <- jags.model("MFweibull.txt", 
                       data = jagsdata, 
                       inits = jagsinits, 
                       n.chains = 1)
    update(jags, 1000)
    samples <- coda.samples(jags, c("beta0", "beta.trt","alpha"), 10000)
    
    # Posterior distribution for log HR (= beta.trt*alpha)
    post_sample <- samples[[1]][, "beta.trt"]*samples[[1]][, "alpha"]
    
    # Calculate posterior probabilities 
    p_any_benefit <- mean(post_sample > 0)      # P(log HR > 0)
    p_true_benefit <- mean(post_sample > beta)  # P(log HR > beta)
    p_90 <- p_any_benefit > 0.90                # P(log HR > 0) > 0.90
    p_true_90 <- p_true_benefit > 0.90
    
    # Extract summary statistics for log HR
    return(c(
      posterior_mean = mean(post_sample), # Mean for log HR
      lower_ci = quantile(post_sample, probs = 0.025), # 2.5% quantiles for log HR
      upper_ci = quantile(post_sample, probs = 0.975), # 97.5% quantiles for log HR
      prob_any_benefit = p_any_benefit, # P(HR > 1)
      power_90 = p_90,                  # P(HR > 1) > 0.90
      level_90 = p_true_90              # P(HR > exp(beta)) > 0.90
    ))
  }, error = function(e) {
    return(rep(NA, 6))
  })
}
```
### **Single Simulation Example**
```{r setup, echo = FALSE, message = FALSE}
library(rjags)

set.seed(2)

n <- 160          # Sample size
beta <- log(1.6)  # log HR (projected true effect)
shape1 <- 0.9    # Shape parameter for primary event 
scale1 <- 4   # Scale parameter for primary event
scale2 <- 21  # Scale parameter for competing event

# Generate treatment assignment
tx <- rbinom(n, 1, 0.5)

# Generate primary event times 
event <- rweibull(n, shape = shape1, scale = (scale1 * exp(-beta * tx))^(1 / shape1))
event <- pmin(event, 28)  # Censor at 21 days 
  
# Generate competing event times 
competing <- rexp(n, rate = 1/scale2)
competing <- pmin(competing, 28)  # Censor at 21 days
  
# Determine event status 
Status <- ifelse(event < competing, 1, 0)
  
simdata <- data.frame(event=event, competing=competing, Status=Status, tx=tx)
  
# set up initial values for MCMC sampling
t<-simdata$event
is.na(t)<-simdata$Status==0
tinits1<- simdata$event+3
is.na(tinits1)<-simdata$Status==1
  
# Visualize survival and censoring distributions
## Split data by treatment group
event0 <- event[tx == 0]
event1 <- event[tx == 1]
competing0 <- competing[tx == 0]
competing1 <- competing[tx == 1]

# Summary statistics function
summarize <- function(x) {
  c(
    Mean = round(mean(x), 1),
    Median = round(median(x), 1),
    IQR = paste0(round(quantile(x, 0.25), 1), "–", round(quantile(x, 0.75), 1))
  )
}

# Create a summary table
summary_df <- data.frame(
  Group = c("Primary Event (Control Group)", "Primary Event (Treatment Group)",
            "Competing Event (Control Group)", "Competing Event (Treatment Group)"),
  rbind(summarize(event0), summarize(event1),
        summarize(competing0), summarize(competing1))
)

# Print summary table
knitr::kable(summary_df, caption = "Summary Statistics of Simulated Times")

# Plot density distributions
plot(density(event0), col = "blue", lwd = 2, xlim = c(0, 30), ylim = c(0, 0.2),
     main = "Simulated Event and Competing Time by Treatment",
     xlab = "Time (days)", ylab = "Density")

lines(density(event1), col = "darkblue", lwd = 2, lty = 2)
lines(density(competing0), col = "red", lwd = 2)
lines(density(competing1), col = "darkred", lwd = 2, lty = 2)

legend("topright",
       legend = c("Primary Event (Tx=0)", "Primary Event (Tx=1)",
                  "Competing Event (Tx=0)", "Competing Event (Tx=1)"),
       col = c("blue", "darkblue", "red", "darkred"),
       lty = c(1, 2, 1, 2),
       lwd = 2,
       bty = "n")
```


```{r, include=FALSE}

# ================================
# --- Plot True Weibull Densities
# ================================

tt <- seq(1e-3, 90, length.out = 300)

scale1_ctrl <- (scale1 * exp(-beta * 0))^(1 / shape1)
scale1_trt <- (scale1 * exp(-beta * 1))^(1 / shape1)

dens_event_tx0 <- dweibull(tt, shape = shape1, scale = scale1_ctrl)
dens_event_tx1 <- dweibull(tt, shape = shape1, scale = scale1_trt)
dens_competing <- dexp(tt, rate = 1/scale2)

# Defensive max() in case of NaNs or all-zero densities
ymax <- max(dens_event_tx0, dens_event_tx1, dens_competing, na.rm = TRUE)

plot(tt, dens_event_tx0, type = "l", col = "blue", lwd = 2,
     ylim = c(0, ymax+1),
     xlab = "Time (days)", ylab = "Density",
     main = "True Weibull Distributions Used for Simulation")

lines(tt, dens_event_tx1, col = "darkblue", lwd = 2, lty = 2)
lines(tt, dens_competing, col = "red", lwd = 2)

legend("topright",
       legend = c("Primary Event (Tx = 0)", "Primary Event (Tx = 1)", "Competing Event"),
       col = c("blue", "darkblue", "red"),
       lty = c(1, 2, 1),
       lwd = 2,
       cex = 0.7)
```

  
### **Interpretation of Bayesian Results**

- Primary inference is **posterior probability** that high ΔP improves outcome.


  - **Any benefit**: `Pr(HR > 1)`
  
  
In the above single simulated example, 
```{r echo=FALSE}
  # rjags model
  jagsdata<-list(t=t,t.cen=simdata$competing,trt=simdata$tx,is.censored=(1-simdata$Status))
  jagsinits<-list(list(t=tinits1,beta0=-1.6))
  jags<-jags.model("MFweibull.txt",data=jagsdata,inits=jagsinits,n.chains=1)
  update(jags,1000)
  
  # --- Draw posterior samples ---
samples <- coda.samples(jags, c("beta0", "beta.trt", "alpha"), n.iter = 10000)

# log HR = log h_T(t)/h_C(t) = beta.trt*alpha
post_sample <- samples[[1]][, "beta.trt"]*samples[[1]][, "alpha"]

cat("Pr(HR > 1) =", round(mean(post_sample > 0), 3))

# --- Summary statistics ---
post_summary <- summary(post_sample)
mean_logHR <- round(mean(post_sample), 3)
ci_logHR <- round(quantile(post_sample, probs = c(0.025, 0.975)), 3)

# --- Plot Posterior Distribution of log(HR) ---
plot(density(post_sample), 
     main = "Posterior Distribution of log(HR)",
     xlab = "log(HR)",
     ylab = "Density",
     lwd = 2,
     col = "darkgreen")

abline(v = 0, col = "red", lty = 2, lwd = 2)          # No effect line
abline(v = log(1.6), col = "blue", lty = 2, lwd = 2)  # True effect

# Add mean and 95% CI for log HR
abline(v = mean_logHR, col = "darkgreen", lty = 1, lwd = 2)
abline(v = ci_logHR[1], col = "gray50", lty = 3)
abline(v = ci_logHR[2], col = "gray50", lty = 3)

# Add legend
legend("topright", 
       legend = c("No Effect (log HR = 0)", 
                  "True log HR = log(1.6)",
                  paste("Posterior Mean =", mean_logHR),
                  paste0("95% CI = [", ci_logHR[1], ", ", ci_logHR[2], "]")),
       col = c("red", "blue", "darkgreen", "gray50"),
       lty = c(2, 2, 1, 3),
       lwd = c(2, 2, 2, 1),
       bty = "n",
       cex = 0.9)
```

- Bayesian Decision Thresholds: We calculate `P(HR > 1)` and declare "success" if this probability exceeds a threshold (e.g., 90%). 
  
- Interpretation: If `Pr(HR > 1) > 0.90`, this is interpreted as a 90% probability that high ΔP is beneficial.

- **Power**: The probability that the posterior probability of any treatment benefit exceeds our decision threshold. Specifically, power is calculated as the proportion of simulations where `Pr(HR > 1 | data) > 0.90`.

- **Type I error**: The probability of incorrectly concluding the treatment effect exceeds a specified threshold when the true effect equals that threshold. For a given true `HR = c`, Type I error is the proportion of simulations where `Pr(HR > c | data) > 0.90`.


## **Sample Size Justification**

Table 2 presents power and Type I error rates across a range of true hazard ratios, based on 1,000 simulations with 160 subjects. The Type I error rate remains consistently controlled below our acceptable level of 0.10 across all scenarios, ranging from 0.072 to 0.095.
 
> *"Using a projected true HR of 1.6 favoring high ΔP, 160 subjects would provide 92.7% power to detect a > 90% probability of any benefit (HR > 1) with high ΔP at a type I error rate of 0.072 assuming a minimally informative prior centered at HR = 1 with precision = 0.2 on the log scale."*

```{r echo=FALSE}
table_data <- data.frame(
  HR = c(1.0, 1.2, 1.4, 1.6),
  Power = c(0.095, 0.399, 0.726, 0.927),
  Level = c(0.095, 0.086, 0.079, 0.072)
)

# View the table
knitr::kable(table_data,
             col.names = c("True HR", "Power", "Type I Error"),
             caption = "Power and Type I Error by True Hazard Ratio")
```
