---
title: "PARMA — Bayesian power for a Weibull time-to-event endpoint"
subtitle: "Time to resolution of hypoxemia, high vs. low driving pressure in pediatric ARDS"
author: "Zhe Chen, with Nadir Yehya"
date: "September 2025"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 4.5)
library(future.apply)
library(rjags)
library(survival)
library(knitr)
library(ggplot2)
plan(multisession, workers = max(1, parallel::detectCores() - 1))
```

Self-contained — the JAGS model is written out below and nothing is sourced
from elsewhere. Copy this folder to start a new time-to-event analysis. Needs
JAGS installed (`brew install jags`).

## The question

PARMA randomises **160** children with ARDS to a high versus low driving
pressure ventilation strategy. The primary endpoint is **time to resolution of
hypoxemia** — alive with PaO₂/FiO₂ > 300 on two measurements four hours apart —
followed for **28 days**. Death acts as a **competing event**.

Because the event is a *good* one, a hazard ratio **greater than 1** is benefit:
the treated arm resolves sooner.

Inference is Bayesian, so the design question is not "will p < 0.05" but:

> How often would this trial produce a posterior probability of benefit,
> `Pr(HR > 1 | data)`, above 0.90?

That frequency **is** the power.

## Assumptions

```{r params}
n_total    <- 160
p_treat    <- 0.5
true_hr    <- 1.6
beta_true  <- log(true_hr)

shape      <- 0.9      # Weibull shape, primary event
scale_ctrl <- 4        # Weibull scale, control arm
mean_comp  <- 21       # mean time to the competing event (death), exponential
horizon    <- 28       # administrative censoring, days

threshold  <- 0.90     # posterior probability bar
prior_trt_prec <- 0.2  # minimally informative; SD ~2.24 on the log-HR scale
prior_b0_prec  <- 0.5
prior_a_shape  <- 1.1
prior_a_rate   <- 1.1

n_iter <- 5000
seed   <- 2026
```

## The parameterisation, written down once

R's `rweibull(shape = a, scale = b)` has $S(t) = \exp\{-(t/b)^a\}$.
JAGS's `dweib(alpha, lambda)` has $S(t) = \exp\{-\lambda t^{\alpha}\}$.
They correspond with $\alpha = a$ and $\lambda = b^{-a}$, so proportional
hazards with log hazard ratio $\beta$ requires

$$b_{\text{trt}} = b_{\text{ctrl}} \cdot e^{-\beta / a}$$

and the baseline intercept on the JAGS scale is
$\beta_0 = -a \log b_{\text{ctrl}}$. This is the one place where a slip would
leave the simulation running perfectly while powering the wrong effect, so it
lives in a single function.

```{r param-helpers}
## ---------------------------------------------------------------------------
## scale_for_arm() — apply the treatment effect to the Weibull scale
##
## The single most important line in this document. Given the control arm's
## scale and a target log hazard ratio, returns the scale the treated arm must
## have for the hazard ratio to come out as intended.
##
## The division by `shape` is the part people get wrong. Writing
## exp(-beta * trt) instead would produce a hazard ratio of exp(beta * shape),
## so with shape = 0.9 you would silently power for HR = 1.53 while believing
## you had specified 1.6. The DGP check further down exists to catch exactly
## this, by fitting a Cox model to a huge simulated trial and confirming the
## recovered log HR matches what was asked for.
## ---------------------------------------------------------------------------
scale_for_arm <- function(scale_ctrl, shape, beta, trt) {
  scale_ctrl * exp(-beta * trt / shape)
}

# The baseline intercept on the JAGS scale implied by the control arm above.
# Used to centre the prior on beta0 honestly, rather than at a round number
# that may not correspond to the data being simulated.
beta0_true <- -shape * log(scale_ctrl)

## ---------------------------------------------------------------------------
## arm_summary() — mean and median survival time for a Weibull(shape, scale)
##
## Purely for reporting: converts the abstract scale parameter into days, so
## the table below can be checked against how the design was described in
## words. Standard results — the mean is b·Γ(1 + 1/a), the median b·(ln 2)^(1/a).
## ---------------------------------------------------------------------------
arm_summary <- function(b, a) c(mean = b * gamma(1 + 1 / a),
                                median = b * log(2)^(1 / a))
ctrl_s <- arm_summary(scale_ctrl, shape)
trt_s  <- arm_summary(scale_for_arm(scale_ctrl, shape, beta_true, 1), shape)
```

```{r arms-table, echo = FALSE}
kable(data.frame(
  Arm = c("Control", "High driving pressure"),
  `Weibull scale` = sprintf("%.2f", c(scale_ctrl,
                                      scale_for_arm(scale_ctrl, shape, beta_true, 1))),
  `Mean time to resolution (days)`   = sprintf("%.1f", c(ctrl_s["mean"], trt_s["mean"])),
  `Median time to resolution (days)` = sprintf("%.1f", c(ctrl_s["median"], trt_s["median"])),
  check.names = FALSE
), caption = sprintf("Implied arms. Shape = %.1f, true HR = %.1f. beta0 = %.3f.",
                     shape, true_hr, beta0_true))
```

```{r dgp-note, echo = FALSE, results = "asis"}
cat(sprintf(
"**Worth flagging.** The original PARMA analysis plan describes these as mean
times to resolution of \"approximately 5 days in the control arm and 3 days in
the intervention arm\". The parameters it actually simulates (shape %.1f, scale
%.0f) imply **%.1f** and **%.1f** days. The hazard ratio — which is what the
power calculation depends on — is exactly the intended %.1f either way, so the
reported power is unaffected. But the prose and the code disagree, and a control
scale of about 4.75 rather than 4 would be needed to make the 5-day description
literally true. Worth reconciling before that text is reused.\n",
shape, scale_ctrl, ctrl_s["mean"], trt_s["mean"], true_hr))
```

## The model

```{r jags-model}
jags_model <- "
model {
  for (i in 1:N) {
    # dinterval() handles right censoring: t[i] is NA when censored, and
    # is.censored[i] = 1 tells JAGS the true value exceeds t.cen[i].
    is.censored[i] ~ dinterval(t[i], t.cen[i])
    t[i] ~ dweib(alpha, lambda[i])
    lambda[i] <- exp(beta0 + beta.trt * trt[i])
  }
  beta0    ~ dnorm(prior_b0_mean, prior_b0_prec)
  beta.trt ~ dnorm(0, prior_trt_prec)
  alpha    ~ dgamma(prior_a_shape, prior_a_rate)
}
"
model_file <- tempfile(fileext = ".jags")
cat(jags_model, file = model_file)
```

`N` and every prior hyperparameter arrive as **data**, so this one model serves
every sample size and every prior in the sensitivity analyses below. The
original version hard-coded `for(i in 1:160)`, which meant a separate file per
scenario.

| Parameter | Prior | Rationale |
|---|---|---|
| `beta.trt` | `dnorm(0, 0.2)` | Minimally informative, centred at **no effect** (HR = 1). Precision 0.2 = SD ≈ 2.24 on the log-HR scale. |
| `beta0` | `dnorm(·, 0.5)` | Baseline hazard, moderately informative. |
| `alpha` | `dgamma(1.1, 1.1)` | Weibull shape; mean 1, variance ≈ 0.91. |

Centring `beta.trt` at zero is what makes the power calculation honest — the
prior is not helping the trial find an effect.

## Machinery

```{r helpers}
## ---------------------------------------------------------------------------
## simulate_trial() — generate ONE hypothetical PARMA trial
##
##   n      number randomised
##   beta   true log hazard ratio (positive = treated resolve sooner = benefit)
##
## Returns one row per participant with:
##   trt     0 = low driving pressure (control), 1 = high
##   time    the observed follow-up time, whatever happened first
##   status  1 = hypoxemia resolved (the event we are powering for)
##           0 = censored, either by death or by reaching day 28
##   t_cen   the time this participant would have been censored at, which
##           JAGS needs separately from `time` (see analyze_bayes below)
## ---------------------------------------------------------------------------
simulate_trial <- function(n, beta) {

  # 1:1 randomisation.
  trt <- rbinom(n, 1, p_treat)

  # Time to resolution, Weibull. scale_for_arm() applies the treatment effect
  # by shifting the scale — see "The parameterisation" section above for why
  # the exponent is -beta/shape and not -beta.
  t_event <- rweibull(n, shape, scale_for_arm(scale_ctrl, shape, beta, trt))

  # Time to death, the competing event. Exponential with a mean of `mean_comp`
  # days. A child who dies never resolves, so this truncates follow-up.
  t_comp <- rexp(n, rate = 1 / mean_comp)

  # Both death and the 28-day horizon stop us observing resolution, so each
  # participant is censored at whichever comes first. Treating the competing
  # event as censoring is what makes this a CAUSE-SPECIFIC hazard model — the
  # estimand is the rate of resolution among those still alive and unresolved,
  # not the cumulative incidence. If the SAP wants a subdistribution hazard,
  # this line and the model both need to change.
  t_cen <- pmin(t_comp, horizon)

  data.frame(
    trt    = trt,
    time   = pmin(t_event, t_cen),               # what we actually observe
    status = as.numeric(t_event <= t_cen),       # did resolution come first?
    t_cen  = t_cen
  )
}

## ---------------------------------------------------------------------------
## analyze_bayes() — fit the Weibull model to one trial and apply the rule
##
##   d          a trial from simulate_trial()
##   beta       the true log HR, used only for the calibration diagnostic below
##   trt_prec   precision of the prior on the treatment effect (varied in the
##              prior sensitivity analysis)
##   b0_mean    mean of the baseline prior (varied to reproduce the memo)
##
## Returns a named vector whose `reject` element is 1 when this trial would
## have been declared a success. Averaging that over trials gives the power.
## ---------------------------------------------------------------------------
analyze_bayes <- function(d, beta, trt_prec = prior_trt_prec,
                          b0_mean = beta0_true) {

  # Data passed to JAGS. Two things are doing real work here:
  #
  #  * `t` is set to NA wherever the participant was censored. JAGS treats NA
  #    as a quantity to be sampled, and the dinterval() line in the model
  #    constrains it to lie above t.cen. That is how censored observations
  #    contribute information without us pretending to know their event time.
  #
  #  * N and every prior hyperparameter go in as DATA rather than being
  #    hard-coded in the model string, which is why one model serves every
  #    sample size and every prior in the sensitivity analyses below.
  jdata <- list(
    N           = nrow(d),
    t           = ifelse(d$status == 1, d$time, NA_real_),
    t.cen       = d$t_cen,
    trt         = d$trt,
    is.censored = 1 - d$status,
    prior_b0_mean  = b0_mean,     prior_b0_prec = prior_b0_prec,
    prior_trt_prec = trt_prec,
    prior_a_shape  = prior_a_shape, prior_a_rate = prior_a_rate
  )

  # Initial values. The censored event times must START above their censoring
  # time or dinterval() is inconsistent at iteration zero and JAGS refuses to
  # initialise — hence t_cen + 1. Uncensored entries are NA because they are
  # observed data, not parameters.
  jinits <- list(list(t = ifelse(d$status == 0, d$t_cen + 1, NA_real_),
                      beta0 = b0_mean, beta.trt = 0))

  # Compile, burn in, then draw the posterior. n.adapt tunes the samplers,
  # update() discards the burn-in, coda.samples() collects the kept draws.
  jm <- jags.model(model_file, jdata, jinits, n.chains = 1,
                   n.adapt = 500, quiet = TRUE)
  update(jm, 1000, progress.bar = "none")
  s <- coda.samples(jm, c("beta0", "beta.trt", "alpha"),
                    n.iter = n_iter, progress.bar = "none")

  # The posterior draws of the log hazard ratio. Everything below is a summary
  # of this one vector.
  post <- as.numeric(s[[1]][, "beta.trt"])
  q    <- unname(quantile(post, c(0.025, 0.975)))

  c(
    # THE DECISION RULE. Not "p < 0.05" — the proportion of posterior draws
    # above zero is Pr(HR > 1 | data), and the trial succeeds when that clears
    # the pre-specified threshold. This is the quantity whose long-run
    # frequency is the Bayesian power.
    reject = as.numeric(mean(post > 0) >= threshold),

    # The same rule applied against the *assumed true* effect rather than the
    # null: how often we would claim more benefit than was actually built in.
    # The original memo reports this as its "type I error". It is a
    # calibration diagnostic, NOT a frequentist type I error rate, and should
    # be described as what it is wherever it appears in a protocol.
    reject_truth = as.numeric(mean(post > beta) >= threshold),

    # A frequentist-style rule on the same posterior, for the comparison with
    # Cox further down: does the 95% credible interval exclude no-effect?
    reject_ci95 = as.numeric(q[1] > 0),

    # Diagnostics, averaged across replicates and reported alongside the power.
    # post_mean should land near the true beta; if it does not, the model and
    # the data-generating process disagree about something.
    prob_benefit = mean(post > 0),
    post_mean    = mean(post),
    event_frac   = mean(d$status)
  )
}

## ---------------------------------------------------------------------------
## run_sim() — repeat a simulated trial many times and summarise
##
##   n_sim    number of hypothetical trials
##   seed     integer fixing the random number streams
##   sim_one  zero-argument function returning one trial's analysed result
##
## Identical in purpose to the version in the PCORI document; the only thing
## that differs between the two analyses is what sim_one() does.
## ---------------------------------------------------------------------------
run_sim <- function(n_sim, seed, sim_one) {

  # Replicates run across CPU cores. Passing an integer to `future.seed` makes
  # future build proper parallel RNG streams, so the result is identical on 1
  # core or 10 — which set.seed() alone would not guarantee under parallelism.
  #
  # tryCatch returning NULL keeps one failed JAGS fit from killing the run.
  out <- future_lapply(seq_len(n_sim),
                       function(i) tryCatch(sim_one(), error = function(e) NULL),
                       future.seed = seed)

  # Count failures rather than dropping them. A model that only converges on
  # the easy scenarios would otherwise look better than it is, so n_failed is
  # reported in the results table and should be checked.
  ok  <- !vapply(out, is.null, logical(1))
  est <- colMeans(do.call(rbind, out[ok]))

  power <- unname(est[["reject"]])
  ci <- wilson_ci(power, sum(ok))
  c(as.list(est), list(mc_se = sqrt(power * (1 - power) / sum(ok)),
                       lo = ci[1], hi = ci[2], n_failed = sum(!ok)))
}

## ---------------------------------------------------------------------------
## wilson_ci() — interval for the SIMULATION's precision, not the trial's
##
## Answers "how much would this power estimate move under a different seed?",
## which is what tells you whether a gap between two scenarios is signal or
## noise. Wilson rather than Wald because Wald collapses to a point at power
## near 0 or 1, and several scenarios here sit above 0.95.
## ---------------------------------------------------------------------------
wilson_ci <- function(p, n, z = 1.96) {
  d      <- 1 + z^2 / n
  centre <- (p + z^2 / (2 * n)) / d
  half   <- z / d * sqrt(p * (1 - p) / n + z^2 / (4 * n^2))
  c(max(0, centre - half), min(1, centre + half))
}
```

## Verify the data-generating model

```{r dgp-check, cache = TRUE}
set.seed(1)
big <- simulate_trial(200000, beta_true)
cf  <- summary(coxph(Surv(time, status) ~ trt, data = big))$coefficients
```

```{r dgp-report, echo = FALSE, results = "asis"}
cat(sprintf(
"True log HR **%.4f**. Cox estimate on n = 200,000: **%.4f** (SE %.4f). Event
fraction by day %g: %.3f. The simulator produces the hazard ratio it was asked
for.\n",
beta_true, cf["trt", "coef"], cf["trt", "se(coef)"], horizon, mean(big$status)))
```

```{r km-plot, echo = FALSE, fig.height = 4}
set.seed(2)
one <- simulate_trial(n_total, beta_true)
plot(survfit(Surv(time, status) ~ trt, data = one),
     col = c("#4C6EF5", "#E8590C"), lwd = 2, mark.time = TRUE,
     xlab = "Days since randomisation",
     ylab = "Proportion with unresolved hypoxemia")
legend("topright", c("Low driving pressure (control)", "High driving pressure"),
       col = c("#4C6EF5", "#E8590C"), lwd = 2, bty = "n")
title(sprintf("One simulated PARMA trial (n = %d)", n_total))
```

## Power

```{r power-main, cache = TRUE}
main <- run_sim(1000, seed,
                function() analyze_bayes(simulate_trial(n_total, beta_true), beta_true))
```

```{r power-table, echo = FALSE}
kable(data.frame(
  Quantity = c("Power: Pr(HR > 1) >= 0.90",
               "Pr(HR > 1) >= 0.90 against the *assumed true* HR of 1.6",
               "95% credible interval excludes HR = 1",
               "Mean posterior Pr(HR > 1)",
               sprintf("Mean posterior log HR (truth = %.3f)", beta_true),
               "Mean event fraction by day 28",
               "Failed fits"),
  Value = c(sprintf("%.3f (MC SE %.3f)", main$reject, main$mc_se),
            sprintf("%.3f", main$reject_truth),
            sprintf("%.3f", main$reject_ci95),
            sprintf("%.3f", main$prob_benefit),
            sprintf("%.3f", main$post_mean),
            sprintf("%.3f", main$event_frac),
            sprintf("%d", main$n_failed)),
  check.names = FALSE
), caption = "1000 simulated trials, n = 160, true HR = 1.6.")
```

The second row is what the original memo reported as a "type I error rate". It
is the same decision rule applied against the *assumed true* effect rather than
the null — how often the trial would claim more benefit than was actually built
in. Useful as a calibration diagnostic, but it is **not** a frequentist type I
error rate and should be described as what it is in a protocol.

### Reproducing the original memo

The submission quoted 93% power at a type I error of 0.101, using a baseline
prior fixed at `dnorm(-1.6, 0.5)` rather than centred on the value the
data-generating control arm implies.

```{r repro, cache = TRUE}
repro_memo <- run_sim(1000, seed, function()
  analyze_bayes(simulate_trial(n_total, beta_true), beta_true, b0_mean = -1.6))
```

```{r repro-table, echo = FALSE}
kable(data.frame(
  Configuration = c("This version (beta0 centred on the DGP)",
                    "Original memo (beta0 ~ dnorm(-1.6, 0.5))"),
  Power = sprintf("%.3f (SE %.3f)", c(main$reject, repro_memo$reject),
                  c(main$mc_se, repro_memo$mc_se)),
  `"Type I error"` = sprintf("%.3f", c(main$reject_truth, repro_memo$reject_truth)),
  check.names = FALSE
), caption = "Memo quoted 93% power and type I error 0.101.")
```

```{r repro-read, echo = FALSE, results = "asis"}
cat(sprintf(
"The headline reproduces: **%.3f** against the quoted 93%%, well inside Monte
Carlo error. The two prior configurations are also indistinguishable from each
other, which says the result does not depend on how the baseline prior was
centred.

The \"type I error\" column comes out at %.3f rather than the quoted 0.101. That
statistic is noisier than the power — it counts how often the posterior clears
the bar against the *assumed true* effect, which is an event of probability
around 0.1, so its Monte Carlo SE at 1000 replicates is about %.3f. The gap is
roughly %.1f standard errors and not evidence of a discrepancy, but it is also
not a number to quote to three decimal places from a single run.\n",
repro_memo$reject, repro_memo$reject_truth,
sqrt(repro_memo$reject_truth * (1 - repro_memo$reject_truth) / 1000),
abs(repro_memo$reject_truth - 0.101) /
  sqrt(repro_memo$reject_truth * (1 - repro_memo$reject_truth) / 1000)))
```

## Sensitivity: sample size and effect size

```{r grid, cache = TRUE}
grid <- expand.grid(n = c(120, 140, 160, 180, 200, 240), hr = c(1.4, 1.6, 1.8))
grid_res <- do.call(rbind, lapply(seq_len(nrow(grid)), function(i) {
  b <- log(grid$hr[i])
  r <- run_sim(400, seed + 10 * i,
               function() analyze_bayes(simulate_trial(grid$n[i], b), b))
  data.frame(power = r$reject, lo = r$lo, hi = r$hi)
}))
grid <- cbind(grid, grid_res)
```

```{r grid-plot, echo = FALSE}
ggplot(grid, aes(n, power, colour = factor(hr))) +
  geom_hline(yintercept = c(0.80, 0.90), linetype = 2, colour = "grey50") +
  geom_ribbon(aes(ymin = lo, ymax = hi, fill = factor(hr)), alpha = 0.15, colour = NA) +
  geom_line(linewidth = 0.8) + geom_point() +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = "Total sample size", y = "Pr( Pr(HR > 1 | data) >= 0.90 )",
       colour = "True HR", fill = "True HR",
       caption = "400 simulations per point; ribbon = Monte Carlo 95% interval.") +
  theme_minimal(base_size = 11)
```

```{r grid-table, echo = FALSE}
w <- reshape(data.frame(n = grid$n, HR = sprintf("HR = %.1f", grid$hr),
                        power = sprintf("%.2f", grid$power)),
             idvar = "n", timevar = "HR", direction = "wide")
names(w) <- sub("^power\\.", "", names(w))
kable(w, row.names = FALSE, caption = "Bayesian power by sample size and true HR.")
```

## Sensitivity: the prior on the treatment effect

The assumption a reviewer will press on. Precision 0.2 is the pre-specified
minimally informative choice; the others bracket it.

```{r prior-sens, cache = TRUE}
precisions <- c(0.05, 0.2, 1, 4)
prior_res <- do.call(rbind, lapply(seq_along(precisions), function(i) {
  r <- run_sim(400, seed + 500 + i, function()
    analyze_bayes(simulate_trial(n_total, beta_true), beta_true,
                  trt_prec = precisions[i]))
  data.frame(precision = precisions[i], sd = 1 / sqrt(precisions[i]),
             power = r$reject, mc_se = r$mc_se, post_mean = r$post_mean)
}))
```

```{r prior-table, echo = FALSE}
kable(data.frame(
  `Prior precision` = prior_res$precision,
  `Prior SD (log HR)` = sprintf("%.2f", prior_res$sd),
  Power = sprintf("%.3f", prior_res$power),
  `MC SE` = sprintf("%.3f", prior_res$mc_se),
  `Mean posterior log HR` = sprintf("%.3f", prior_res$post_mean),
  `Shrinkage vs. truth` = sprintf("%+.3f", prior_res$post_mean - beta_true),
  check.names = FALSE
), caption = "Effect of the treatment-effect prior. All centred at HR = 1.")
```

```{r prior-read, echo = FALSE, results = "asis"}
tight <- which.max(prior_res$precision)
cat(sprintf(
"Read this as a *negative* result, which is the useful kind here. Power spans
only %.3f across all four settings — from a prior SD of %.2f down to %.2f on the
log-HR scale — against a Monte Carlo SE of about %.3f on each estimate. They are
statistically indistinguishable, and the ordering between them is noise rather
than signal. At n = %d this design is simply not sensitive to the
treatment-effect prior.

Where the tightest prior *does* show up is in the estimate rather than the
decision: at precision %.0f (prior SD %.2f) the posterior mean falls from the
true %.3f to %.3f, a shrinkage of %.3f toward the null the prior is centred on.
That is the expected behaviour and worth showing, but it does not translate into
a power penalty here — the posterior still clears the %.2f bar about as often.
A prior that tight is comparable in width to the effect being estimated, so it
would be a strange thing to pre-specify; it is here to locate the boundary, not
to recommend it.\n",
max(prior_res$power) - min(prior_res$power),
prior_res$sd[which.min(prior_res$precision)], prior_res$sd[tight],
mean(prior_res$mc_se), n_total,
prior_res$precision[tight], prior_res$sd[tight],
beta_true, prior_res$post_mean[tight],
prior_res$post_mean[tight] - beta_true, threshold))
```

## Frequentist comparator on identical data

A Cox model on the same simulated trials, so any difference is attributable to
the analysis and not to differently simulated data.

```{r cox, cache = TRUE}
cox_res <- run_sim(1000, seed + 900, function() {
  d <- simulate_trial(n_total, beta_true)
  fit <- summary(coxph(Surv(time, status) ~ trt, data = d))$coefficients
  c(reject = as.numeric(fit["trt", "Pr(>|z|)"] < 0.05))
})
```

```{r cox-report, echo = FALSE, results = "asis"}
cat(sprintf(
"At n = %d the Bayesian 95%% credible interval excludes the null **%.3f** of the
time and the Cox test rejects at p < 0.05 **%.3f** of the time. That agreement
is the expected result and worth stating: with a treatment-effect prior this
weak, a 95%% credible interval behaves almost exactly like a two-sided 5%% test,
so the Bayesian machinery is not adding power through the back door. It is a
useful consistency check on both implementations.

The trial's actual decision rule — a one-sided posterior probability above
%.2f — gives **%.3f**, roughly %.0f points more. That gap is a deliberate choice
about how much evidence is required, not a modelling artefact, and should be
presented to reviewers and to the DSMB as such rather than compared silently
against a conventional 5%% test.\n",
n_total, main$reject_ci95, cox_res$reject, threshold, main$reject,
100 * (main$reject - cox_res$reject)))
```

## Assumptions and limitations

- Control-arm Weibull parameters are calibrated to empirical CHOP data; the
  shape of 0.9 is estimated, not assumed exactly.
- Death is treated as censoring for the cause-specific hazard of resolution.
  That matches a cause-specific estimand, not a subdistribution one.
- No dropout beyond the competing event and the 28-day horizon.
- The simulation fits the same Weibull PH model that the SAP specifies. If the
  analysis changes, re-run.

## Notes for reuse

- **Bad event** (death, readmission): benefit is HR < 1. Flip `beta_true` and
  read the rule as `Pr(HR < 1) >= threshold`.
- **Weibull competing event**: swap `rexp()` for `rweibull()` in
  `simulate_trial()`.
- **Exponential baseline**: fix `alpha` at 1 in the model string.
- **Runtime**: about 13,000 JAGS fits, roughly 15 minutes cold on 8 cores. Keep
  `cache = TRUE` on the compute chunks and `kable()` in separate chunks, so
  editing a table does not re-run the simulation.

## Reproducibility

```{r session, echo = FALSE}
sessionInfo()$R.version$version.string
```

Seeds are fixed in every call above.
