---
title: "STUDY — Bayesian power for a time-to-event outcome"
subtitle: "Weibull proportional-hazards model, fitted in JAGS"
author: "Your name"
date: "Month Year"
output:
  html_document:
    toc: true
    toc_float: true
---

<!--
TEMPLATE — parallel-arm trial, time-to-event outcome, Bayesian analysis.

Copy this file to analyses/YYYY-MM_your-slug/analysis.Rmd, then work through it
top to bottom replacing the placeholder numbers. It runs as-is — knit it once
before changing anything and confirm you get output. Takes a couple of minutes.

Everything is in this file, including the JAGS model. Nothing is source()d.

Needs JAGS installed:  brew install jags
When you're done, add a row to the index table in ../../README.md.
-->

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 4.5)
library(future.apply)
library(rjags)
library(survival)
library(knitr)
library(ggplot2)
plan(multisession, workers = max(1, parallel::detectCores() - 1))
```

## The question

Replace this with the design. For a Bayesian trial the decision rule matters as
much as the effect size, so state it explicitly:

- **Design** — arms, randomisation, allocation ratio, follow-up length.
- **Primary endpoint** — and whether the event is *good* (recovery, discharge)
  or *bad* (death, readmission). This fixes which direction of the hazard ratio
  counts as benefit.
- **Competing events**, if any, and how they are handled.
- **Target effect**, as a hazard ratio, and where it came from.
- **Decision rule** — the posterior probability and the bar it must clear.
- **Priors**, and which one a reviewer will press on.

The design question here is not "will p < 0.05". It is:

> How often would this trial produce a posterior probability of benefit above
> the pre-specified threshold?

That frequency **is** the power.

## Assumptions

```{r params}
n_total    <- 160        # total randomised
p_treat    <- 0.5        # randomisation probability
true_hr    <- 1.6        # hazard ratio worth detecting
beta_true  <- log(true_hr)

shape      <- 0.9        # Weibull shape for the primary event
scale_ctrl <- 4          # Weibull scale, control arm
mean_comp  <- 21         # mean time to the competing event (exponential)
horizon    <- 28         # administrative censoring, days

threshold  <- 0.90       # posterior probability bar for declaring success
prior_trt_prec <- 0.2    # precision on beta.trt; 0.2 => SD ~2.24 on log-HR scale
prior_b0_prec  <- 0.5
prior_a_shape  <- 1.1
prior_a_rate   <- 1.1

n_sim  <- 400            # raise for the number that goes in the application
n_iter <- 5000
seed   <- 2026
```

**The event here is a good one**, so a hazard ratio **greater than 1** is
benefit — the treated arm reaches the event sooner. If your endpoint is a bad
one, flip the sign of `beta_true` and read the decision rule as
`Pr(HR < 1) >= threshold`.

## The parameterisation, written down once

This is the step that quietly breaks these simulations, so it gets stated
explicitly rather than buried.

R's `rweibull(shape = a, scale = b)` has $S(t) = \exp\{-(t/b)^a\}$.
JAGS's `dweib(alpha, lambda)` has $S(t) = \exp\{-\lambda t^{\alpha}\}$.
They correspond with $\alpha = a$ and $\lambda = b^{-a}$. Proportional hazards
with log hazard ratio $\beta$ therefore needs

$$b_{\text{trt}} = b_{\text{ctrl}} \cdot e^{-\beta / a}$$

Get that exponent wrong and the simulation runs perfectly while powering a
different effect than you think. The consequence for the baseline intercept on
the JAGS scale is $\beta_0 = -a \log b_{\text{ctrl}}$, which is what the
baseline prior should be centred on.

```{r param-helpers}
scale_for_arm <- function(scale_ctrl, shape, beta, trt) {
  scale_ctrl * exp(-beta * trt / shape)
}
beta0_true <- -shape * log(scale_ctrl)

# Implied control and treated arms, for reporting.
arm_summary <- function(b, a) {
  c(mean = b * gamma(1 + 1 / a), median = b * log(2)^(1 / a))
}
ctrl_s <- arm_summary(scale_ctrl, shape)
trt_s  <- arm_summary(scale_for_arm(scale_ctrl, shape, beta_true, 1), shape)
```

```{r arms-table, echo = FALSE}
kable(data.frame(
  Arm = c("Control", "Treatment"),
  `Mean time to event` = sprintf("%.1f", c(ctrl_s["mean"], trt_s["mean"])),
  `Median time to event` = sprintf("%.1f", c(ctrl_s["median"], trt_s["median"])),
  check.names = FALSE
), caption = sprintf("Implied arms. Shape = %.2g, true HR = %.2g.", shape, true_hr))
```

Check these against however the design was described in words. If the protocol
says "a mean of 5 days in the control arm" and this table says 4.2, the prose
and the code disagree and one of them needs fixing.

## The model

Written here as a string and handed to JAGS from a temporary file, so this
document carries its own model rather than depending on a file elsewhere.

```{r jags-model}
jags_model <- "
model {
  for (i in 1:N) {
    # dinterval() handles right censoring: t[i] is NA when censored, and
    # is.censored[i] = 1 tells JAGS the true value exceeds t.cen[i].
    is.censored[i] ~ dinterval(t[i], t.cen[i])
    t[i] ~ dweib(alpha, lambda[i])
    lambda[i] <- exp(beta0 + beta.trt * trt[i])
  }
  beta0    ~ dnorm(prior_b0_mean, prior_b0_prec)
  beta.trt ~ dnorm(0, prior_trt_prec)
  alpha    ~ dgamma(prior_a_shape, prior_a_rate)
}
"
model_file <- tempfile(fileext = ".jags")
cat(jags_model, file = model_file)
```

`N` and every prior hyperparameter arrive as **data**, not as literals, so this
one model serves every sample size and every prior in the sensitivity analyses
below without editing.

| Parameter | Prior | Why |
|---|---|---|
| `beta.trt` | `dnorm(0, 0.2)` | Centred at **no effect**. This is what keeps the power calculation honest — the prior is not helping the trial find an effect. |
| `beta0` | `dnorm(beta0_true, 0.5)` | Baseline hazard, centred on the value the data-generating control arm implies. |
| `alpha` | `dgamma(1.1, 1.1)` | Weibull shape; mean 1. |

## Machinery

```{r helpers}
# Simulate one trial: Weibull event times, a competing event, and an
# administrative horizon. Both the competing event and the horizon act as
# right censoring for the cause-specific hazard of the primary event.
simulate_trial <- function(n, beta) {
  trt <- rbinom(n, 1, p_treat)
  t_event <- rweibull(n, shape, scale_for_arm(scale_ctrl, shape, beta, trt))
  t_comp  <- rexp(n, rate = 1 / mean_comp)
  t_cen   <- pmin(t_comp, horizon)
  data.frame(trt = trt,
             time   = pmin(t_event, t_cen),
             status = as.numeric(t_event <= t_cen),
             t_cen  = t_cen)
}

# Fit the Bayesian model to one trial and apply the decision rule.
analyze_bayes <- function(d, beta, trt_prec = prior_trt_prec) {
  jdata <- list(
    N = nrow(d),
    t = ifelse(d$status == 1, d$time, NA_real_),   # NA where censored
    t.cen = d$t_cen, trt = d$trt,
    is.censored = 1 - d$status,
    prior_b0_mean = beta0_true, prior_b0_prec = prior_b0_prec,
    prior_trt_prec = trt_prec,
    prior_a_shape = prior_a_shape, prior_a_rate = prior_a_rate
  )
  # Censored event times must be initialised above their censoring time.
  jinits <- list(list(t = ifelse(d$status == 0, d$t_cen + 1, NA_real_),
                      beta0 = beta0_true, beta.trt = 0))

  jm <- jags.model(model_file, jdata, jinits, n.chains = 1,
                   n.adapt = 500, quiet = TRUE)
  update(jm, 1000, progress.bar = "none")
  s <- coda.samples(jm, c("beta0", "beta.trt", "alpha"),
                    n.iter = n_iter, progress.bar = "none")

  post <- as.numeric(s[[1]][, "beta.trt"])
  q <- unname(quantile(post, c(0.025, 0.975)))

  c(reject       = as.numeric(mean(post > 0) >= threshold),
    reject_ci95  = as.numeric(q[1] > 0),   # 95% CrI excludes the null
    prob_benefit = mean(post > 0),
    post_mean    = mean(post),
    event_frac   = mean(d$status))
}

run_sim <- function(n_sim, seed, sim_one) {
  out <- future_lapply(seq_len(n_sim),
                       function(i) tryCatch(sim_one(), error = function(e) NULL),
                       future.seed = seed)
  ok <- !vapply(out, is.null, logical(1))
  est <- colMeans(do.call(rbind, out[ok]))
  power <- unname(est[["reject"]])
  ci <- wilson_ci(power, sum(ok))
  c(as.list(est), list(mc_se = sqrt(power * (1 - power) / sum(ok)),
                       lo = ci[1], hi = ci[2], n_failed = sum(!ok)))
}

wilson_ci <- function(p, n, z = 1.96) {
  d <- 1 + z^2 / n
  centre <- (p + z^2 / (2 * n)) / d
  half   <- z / d * sqrt(p * (1 - p) / n + z^2 / (4 * n^2))
  c(max(0, centre - half), min(1, centre + half))
}
```

## Verify the data-generating model

Before powering anything, confirm the simulator produces the hazard ratio it was
asked for. Fit a Cox model to one very large simulated trial — if the
parameterisation above is wrong, this is where it shows.

```{r dgp-check, cache = TRUE}
set.seed(1)
big <- simulate_trial(200000, beta_true)
cf  <- summary(coxph(Surv(time, status) ~ trt, data = big))$coefficients
```

```{r dgp-report, echo = FALSE, results = "asis"}
cat(sprintf(
"True log HR **%.4f**. Cox estimate on n = 200,000: **%.4f** (SE %.4f).
Event fraction by day %g: %.3f.\n",
beta_true, cf["trt", "coef"], cf["trt", "se(coef)"], horizon, mean(big$status)))
```

```{r km-plot, echo = FALSE, fig.height = 4}
set.seed(2)
one <- simulate_trial(n_total, beta_true)
plot(survfit(Surv(time, status) ~ trt, data = one),
     col = c("#4C6EF5", "#E8590C"), lwd = 2, mark.time = TRUE,
     xlab = "Days since randomisation", ylab = "Proportion without the event")
legend("topright", c("Control", "Treatment"),
       col = c("#4C6EF5", "#E8590C"), lwd = 2, bty = "n")
title(sprintf("One simulated trial (n = %d)", n_total))
```

## Power

```{r power, cache = TRUE}
main <- run_sim(n_sim, seed, function() analyze_bayes(simulate_trial(n_total, beta_true), beta_true))
```

```{r power-table, echo = FALSE}
kable(data.frame(
  Quantity = c(sprintf("Power: Pr(HR > 1) >= %.2f", threshold),
               "95% credible interval excludes HR = 1",
               "Mean posterior Pr(HR > 1)",
               sprintf("Mean posterior log HR (truth = %.3f)", beta_true),
               "Mean event fraction",
               "Failed fits"),
  Value = c(sprintf("%.3f (MC SE %.3f)", main$reject, main$mc_se),
            sprintf("%.3f", main$reject_ci95),
            sprintf("%.3f", main$prob_benefit),
            sprintf("%.3f", main$post_mean),
            sprintf("%.3f", main$event_frac),
            sprintf("%d", main$n_failed)),
  check.names = FALSE
), caption = sprintf("%d simulated trials, n = %d, true HR = %.2g.",
                     n_sim, n_total, true_hr))
```

## Sensitivity: sample size and effect size

```{r sweep, cache = TRUE}
sweep <- expand.grid(n = c(120, 160, 200), hr = c(1.4, 1.6, 1.8))
sweep_res <- do.call(rbind, lapply(seq_len(nrow(sweep)), function(i) {
  b <- log(sweep$hr[i])
  r <- run_sim(200, seed + 10 * i,
               function() analyze_bayes(simulate_trial(sweep$n[i], b), b))
  data.frame(power = r$reject, lo = r$lo, hi = r$hi)
}))
sweep <- cbind(sweep, sweep_res)
```

```{r sweep-plot, echo = FALSE}
ggplot(sweep, aes(n, power, colour = factor(hr))) +
  geom_hline(yintercept = c(0.80, 0.90), linetype = 2, colour = "grey50") +
  geom_ribbon(aes(ymin = lo, ymax = hi, fill = factor(hr)), alpha = 0.15, colour = NA) +
  geom_line() + geom_point() +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = "Total sample size",
       y = sprintf("Pr( Pr(HR > 1 | data) >= %.2f )", threshold),
       colour = "True HR", fill = "True HR",
       caption = "200 simulations per point; ribbon = Monte Carlo 95% interval.") +
  theme_minimal(base_size = 11)
```

## Sensitivity: the prior on the treatment effect

The assumption a reviewer will press on. Report what the run actually shows —
if power is flat across every prior you would defend, that is a useful negative
result and worth stating as one.

```{r prior-sens, cache = TRUE}
precisions <- c(0.05, 0.2, 1, 4)
prior_res <- do.call(rbind, lapply(seq_along(precisions), function(i) {
  r <- run_sim(200, seed + 500 + i,
               function() analyze_bayes(simulate_trial(n_total, beta_true),
                                        beta_true, trt_prec = precisions[i]))
  data.frame(precision = precisions[i], sd = 1 / sqrt(precisions[i]),
             power = r$reject, mc_se = r$mc_se, post_mean = r$post_mean)
}))
```

```{r prior-table, echo = FALSE}
kable(data.frame(
  `Prior precision` = prior_res$precision,
  `Prior SD (log HR)` = sprintf("%.2f", prior_res$sd),
  Power = sprintf("%.3f", prior_res$power),
  `MC SE` = sprintf("%.3f", prior_res$mc_se),
  `Mean posterior log HR` = sprintf("%.3f", prior_res$post_mean),
  check.names = FALSE
), caption = "Effect of the treatment-effect prior. All centred at HR = 1.")
```

## Frequentist comparator

A Cox model on the **same** simulated trials, so any difference is attributable
to the analysis and not to differently simulated data. If the Bayesian 95%
credible interval and a two-sided 5% test disagree badly under a weak prior,
something is wrong with one of the implementations.

```{r cox, cache = TRUE}
cox_res <- run_sim(n_sim, seed + 900, function() {
  d <- simulate_trial(n_total, beta_true)
  fit <- summary(coxph(Surv(time, status) ~ trt, data = d))$coefficients
  c(reject = as.numeric(fit["trt", "Pr(>|z|)"] < 0.05))
})
```

```{r cox-report, echo = FALSE, results = "asis"}
cat(sprintf(
"At n = %d: Bayesian 95%% CrI excludes the null **%.3f** of the time, Cox
p < 0.05 **%.3f** of the time. The trial's actual rule — a one-sided posterior
probability above %.2f — gives **%.3f**.

That last gap is a deliberate choice about how much evidence is required, not a
modelling artefact. Present it to reviewers as such rather than comparing it
silently against a conventional 5%% test.\n",
n_total, main$reject_ci95, cox_res$reject, threshold, main$reject))
```

## Assumptions and limitations

- Where the control-arm survival parameters came from.
- Whether the Weibull shape is supported by data or assumed.
- How the competing event is handled, and whether treating it as censoring
  matches the estimand in the SAP.
- Dropout beyond the competing event and the horizon.
- Whether the analysis model here matches the one that will actually be fitted.

## Adapting this template

- **Bad event** (death, readmission): benefit is HR < 1. Flip `beta_true` and
  read the rule as `Pr(HR < 1) >= threshold`.
- **No competing risk**: drop `t_comp` and set `t_cen <- horizon`.
- **Weibull competing event**: swap `rexp()` for `rweibull()`.
- **Exponential baseline**: fix `alpha` at 1 in the model string.
- **Runtime**: one JAGS fit is ~60 ms, so 1000 replicates is about a minute on 8
  cores. Keep `n_sim` low while exploring and raise it for the final numbers —
  `mc_se` tells you when it is high enough. Keep `cache = TRUE` on the compute
  chunks and keep `kable()` in a separate chunk, so editing a table does not
  re-run the simulation.

## Reproducibility

```{r session, echo = FALSE}
sessionInfo()$R.version$version.string
```

Seeds are fixed in every call above, so re-knitting reproduces these numbers.
