---
title: "STUDY — power for a binary primary outcome"
subtitle: "Parallel-arm randomised trial"
author: "Your name"
date: "Month Year"
output:
  html_document:
    toc: true
    toc_float: true
---

<!--
TEMPLATE — parallel-arm trial, binary outcome.

Copy this file to analyses/YYYY-MM_your-slug/analysis.Rmd, then work through it
top to bottom replacing the placeholder numbers. It runs as-is, so knit it once
before you change anything and confirm you get output.

Everything is in this file. Nothing is source()d from elsewhere, so you can read
it start to finish and copy it somewhere else without dragging a library along.

When you're done, add a row to the index table in ../../README.md.
-->

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 4.5)
library(future.apply)
library(knitr)
library(ggplot2)
plan(multisession, workers = max(1, parallel::detectCores() - 1))
```

## The question

Replace this with the design, in sentences a clinician will read:

- **Design** — how many arms, how randomised, allocation ratio.
- **Primary outcome** — exactly as the protocol will define it.
- **Control event rate**, and where it came from (pilot, registry, prior trial,
  clinical judgement — say which).
- **Effect worth detecting**, and the same question about where it came from.
- **Constraint** — fixed N? fixed sites? fixed recruitment window?
- **Multiplicity** — how many comparisons are confirmatory, and the adjustment.

If a number here is a guess, say so. Those are the ones the sensitivity analysis
below exists to stress.

## Assumptions

Everything the simulation depends on, in one place, so a reader never has to
hunt for a number.

```{r params}
p_control  <- 0.40    # control arm event rate
p_treat    <- 0.55    # treatment arm event rate under the alternative
n_per_arm  <- 200     # participants per arm
alpha      <- 0.05    # family-wise two-sided level
n_compare  <- 1       # confirmatory comparisons; >1 applies Bonferroni
n_sim      <- 4000    # simulation replicates
seed       <- 2026

alpha_test <- alpha / n_compare   # the level each test actually uses
```

## Machinery

Three small helpers. They are the only infrastructure in this document.

```{r helpers}
# One simulated trial: generate, analyse, return a named vector that MUST
# contain `reject`. Anything else it returns is averaged and reported too.
simulate_one <- function(p_control, p_treat, n_per_arm, alpha_test) {
  y0 <- rbinom(n_per_arm, 1, p_control)
  y1 <- rbinom(n_per_arm, 1, p_treat)

  # Uncorrected two-sample test of proportions. A table with no events (or no
  # non-events) in both arms gives 0/0; there is no evidence against the null
  # there, so call it a non-rejection rather than letting NaN through.
  pv <- suppressWarnings(
    prop.test(c(sum(y1), sum(y0)), c(n_per_arm, n_per_arm), correct = FALSE)$p.value
  )
  if (!is.finite(pv)) pv <- 1

  c(reject     = as.numeric(pv < alpha_test),
    risk_diff  = mean(y1) - mean(y0))
}

# Replicate and average. `future.seed` takes an integer, which makes the result
# reproducible no matter how many workers are used.
run_sim <- function(n_sim, seed, sim_one) {
  out <- future_lapply(seq_len(n_sim),
                       function(i) tryCatch(sim_one(), error = function(e) NULL),
                       future.seed = seed)
  ok <- !vapply(out, is.null, logical(1))
  est <- colMeans(do.call(rbind, out[ok]))
  power <- unname(est[["reject"]])
  ci <- wilson_ci(power, sum(ok))
  c(as.list(est),
    list(mc_se = sqrt(power * (1 - power) / sum(ok)),
         lo = ci[1], hi = ci[2], n_failed = sum(!ok)))
}

# Wilson rather than Wald: at power near 0 or 1 the Wald interval collapses to
# a point, which is exactly where the validation below needs it to be sane.
wilson_ci <- function(p, n, z = 1.96) {
  d <- 1 + z^2 / n
  centre <- (p + z^2 / (2 * n)) / d
  half   <- z / d * sqrt(p * (1 - p) / n + z^2 / (4 * n^2))
  c(max(0, centre - half), min(1, centre + half))
}
```

## Validation

Before trusting any power number, check the simulation against a value computed
a different way.

For two proportions we can do better than an approximation: enumerate **every
2×2 table the trial could produce**, and sum the probability of the ones that
reject. That is the exact power of the test being simulated, so agreement is a
real correctness check rather than two approximations agreeing with each other.

```{r exact}
# Exact power of the uncorrected two-proportion test, by enumeration.
# Cost is (n1+1) x (n0+1), so keep it under a few thousand per arm.
power_exact <- function(n1, n0, p1, p0, alpha) {
  x1 <- 0:n1; x0 <- 0:n0
  pbar <- outer(x1, x0, "+") / (n1 + n0)
  se   <- sqrt(pbar * (1 - pbar) * (1 / n1 + 1 / n0))
  z    <- outer(x1 / n1, x0 / n0, "-") / se
  rej  <- is.finite(z) & abs(z) > qnorm(1 - alpha / 2)
  sum(outer(dbinom(x1, n1, p1), dbinom(x0, n0, p0))[rej])
}
```

```{r validate, cache = TRUE}
check <- run_sim(n_sim, seed,
                 function() simulate_one(p_control, p_treat, n_per_arm, alpha_test))
exact <- power_exact(n_per_arm, n_per_arm, p_treat, p_control, alpha_test)
```

```{r validate-report, echo = FALSE, results = "asis"}
cat(sprintf(
"Simulated power **%.4f** (MC SE %.4f, 95%% interval %.4f–%.4f) against an exact
value of **%.4f**. Difference %.4f, which %s inside the Monte Carlo interval.
Failed replicates: %d.\n",
check$reject, check$mc_se, check$lo, check$hi, exact,
abs(check$reject - exact),
if (exact >= check$lo && exact <= check$hi) "is" else "is *not*",
check$n_failed))
```

A note on why this matters: `power.prop.test()` uses a normal approximation and
runs up to about 0.02 conservative at n = 150 per arm. Validating against it
cannot distinguish "the simulation is right" from "both approximations are wrong
in the same direction". Use it for the conventional number a reviewer will
reproduce by hand — not as the thing you check against.

## Type I error

The same machinery with no true effect. A simulation that cannot hold its
nominal level is not measuring power.

```{r null, cache = TRUE}
null <- run_sim(n_sim, seed + 1,
                function() simulate_one(p_control, p_control, n_per_arm, alpha_test))
```

```{r null-report, echo = FALSE, results = "asis"}
cat(sprintf(
"Empirical type I error **%.4f** (95%% interval %.4f–%.4f) against a nominal
%.4f. The chi-square test on discrete data is not exactly its nominal level, so
the exact size — %.4f — is the fairer target.\n",
null$reject, null$lo, null$hi, alpha_test,
power_exact(n_per_arm, n_per_arm, p_control, p_control, alpha_test)))
```

## Required sample size

The conventional table, using the normal approximation because that is what
protocols quote and what a reviewer will check by hand. It is mildly
conservative, so the required *n* errs on the safe side.

```{r required-n}
req_n <- function(p0, p1, alpha, target = 0.80) {
  ceiling(power.prop.test(p1 = p0, p2 = p1, sig.level = alpha,
                          power = target, alternative = "two.sided")$n)
}

controls <- c(0.30, 0.40, 0.50)
effects  <- c(0.05, 0.075, 0.10)
grid <- expand.grid(control = controls, effect = effects)
grid <- grid[order(grid$control, grid$effect), ]
grid$treat <- grid$control + grid$effect
grid$n     <- mapply(req_n, grid$control, grid$treat, alpha_test)

pct <- function(x) paste0(trimws(formatC(x * 100, format = "fg", digits = 3)), "%")

kable(data.frame(
  Control = pct(grid$control),
  Treatment = pct(grid$treat),
  `Difference (pp)` = grid$effect * 100,
  `n per arm` = grid$n,
  check.names = FALSE
), caption = sprintf("Required n per arm for 80%% power at alpha = %.4g.", alpha_test))
```

## Power across scenarios

Vary whatever you were least sure about in "The question". Here it is the
control rate and the effect size, at a fixed sample size.

```{r sweep, cache = TRUE}
sweep <- expand.grid(control = controls, effect = effects)
sweep <- sweep[order(sweep$control, sweep$effect), ]
sweep$treat <- sweep$control + sweep$effect

sweep_res <- do.call(rbind, lapply(seq_len(nrow(sweep)), function(i) {
  r <- run_sim(1000, seed + 100 + i,
               function() simulate_one(sweep$control[i], sweep$treat[i],
                                       n_per_arm, alpha_test))
  data.frame(power = r$reject, lo = r$lo, hi = r$hi, mc_se = r$mc_se)
}))
sweep <- cbind(sweep, sweep_res)
```

```{r sweep-table, echo = FALSE}
kable(data.frame(
  Control = pct(sweep$control),
  Treatment = pct(sweep$treat),
  `Difference (pp)` = sweep$effect * 100,
  Power = sprintf("%.3f", sweep$power),
  `MC SE` = sprintf("%.3f", sweep$mc_se),
  check.names = FALSE
), caption = sprintf("Simulated power at %d per arm.", n_per_arm))
```

```{r sweep-plot, echo = FALSE}
ggplot(sweep, aes(effect * 100, power, colour = factor(control * 100))) +
  geom_hline(yintercept = 0.80, linetype = 2, colour = "grey45") +
  geom_ribbon(aes(ymin = lo, ymax = hi, fill = factor(control * 100)),
              alpha = 0.15, colour = NA) +
  geom_line() + geom_point(size = 1.3) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = "Absolute difference vs. control (percentage points)",
       y = "Power", colour = "Control\nrate (%)", fill = "Control\nrate (%)",
       caption = "Dashed line = 80% power. Ribbon = Monte Carlo 95% interval.") +
  theme_minimal(base_size = 11)
```

## Minimum detectable effect

The single number a grant application usually needs.

```{r mde, echo = FALSE, results = "asis"}
mde <- function(n, p0, alpha, target = 0.80) {
  uniroot(function(d) power.prop.test(n, p0, p0 + d, alpha)$power - target,
          c(0.005, 0.40))$root
}
cat(sprintf(
"> With %d participants per arm, the trial has 80%% power to detect an absolute
difference of **%.1f percentage points** against a %s control event rate
(two-sided alpha = %.4g).\n",
n_per_arm, mde(n_per_arm, p_control, alpha_test) * 100, pct(p_control), alpha_test))
```

## Assumptions and limitations

Write these out — they are what a reviewer probes.

- Where the control event rate came from.
- Dropout and loss to follow-up: in the simulation, or handled by inflating N?
- Whether the analysis simulated here matches the one in the SAP.
- What the simulation does **not** represent — site effects, interim analyses,
  missing outcomes, non-adherence.

## Adapting this template

- **More than two arms.** Generate a vector of arms and test the pair you care
  about. Set `n_compare` to the number of confirmatory comparisons.
- **Subgroup power.** Same code with a smaller `n_per_arm` — if a stratum is a
  third of the trial, run it at a third of the arm size.
- **Unequal allocation.** Give `simulate_one()` separate `n1` / `n0` and pass
  both to `power_exact()`, which already accepts them.
- **A different test.** Swap `prop.test()` for `fisher.test()` or a logistic
  model. Note the exact enumeration above is specific to the uncorrected
  chi-square test, so it stops being the right yardstick if you change this.

## Reproducibility

```{r session, echo = FALSE}
sessionInfo()$R.version$version.string
```

Seeds are fixed in every call above, so re-knitting reproduces these numbers.
