---
title: "PCORI LOI — three-arm pragmatic trial, binary outcome"
subtitle: "Sample size, achieved power at budget, and subgroup power"
author: "Zhe Chen"
date: "September 2026"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 4.5)
library(future.apply)
library(knitr)
library(ggplot2)
plan(multisession, workers = max(1, parallel::detectCores() - 1))
```

Self-contained — nothing is sourced from elsewhere. Copy this folder to start
a new binary-outcome analysis.

## The question

Three arms — one control and two **different** active interventions — with a
fixed budget of **N = 3000**, so 1000 per arm. The clinicians need to read two
things straight off a table:

1. How large does an arm have to be to detect a given improvement over control?
2. At the budget we actually have, what can we detect — overall, and within the
   rural and urban subgroups?

The two active arms are **not** assumed to have the same effect. Every table
varies one arm at a time, so a reader can pick the row matching whatever they
believe about each intervention separately.

### Multiplicity

Two primary comparisons, each active against control → Bonferroni
**α = 0.025**. If the active-versus-active contrast is also treated as
confirmatory, three pairwise comparisons → **α = 0.0167**. Both appear below.
Which one applies is a protocol decision, not a statistical one.

## Assumptions

```{r params}
n_arm      <- 1000          # budget: 3000 total across 3 arms
alpha      <- 0.05
alpha_2    <- alpha / 2     # 0.025  — two primary comparisons
alpha_3    <- alpha / 3     # 0.0167 — three pairwise comparisons
n_sim      <- 20000
seed       <- 2026

# The trial is 1/3 rural, 2/3 urban, so each arm contributes:
n_rural <- round(n_arm / 3)       # 333
n_urban <- round(n_arm * 2 / 3)   # 667

controls <- seq(0.30, 0.50, by = 0.05)
diffs    <- c(0.050, 0.075, 0.100)
```

## Machinery

```{r helpers}
## ---------------------------------------------------------------------------
## simulate_one() — generate and analyse ONE hypothetical trial
##
## This is the heart of the simulation, and the only function you need to
## rewrite if the design changes. Everything else is bookkeeping.
##
##   p_control   true event rate in the control arm
##   p_treat     true event rate in the active arm (the alternative we power for)
##   n_per_arm   participants in each arm
##   alpha_test  the level THIS test uses, already Bonferroni-adjusted
##
## Returns a named vector that must contain `reject` — 1 if this simulated
## trial would have declared a significant difference, 0 if not. Averaging
## `reject` over many trials is the power. Anything else returned here is
## averaged and reported alongside it, which is a convenient way to check the
## simulation is doing what you think (e.g. add `risk_diff` to confirm the
## realised effect matches the one you specified).
## ---------------------------------------------------------------------------
simulate_one <- function(p_control, p_treat, n_per_arm, alpha_test) {

  # Each participant is an independent coin flip at their arm's event rate.
  # This is the data-generating model: no clustering, no covariates, no
  # dropout. If the real trial has any of those, this line is what changes.
  y0 <- rbinom(n_per_arm, 1, p_control)
  y1 <- rbinom(n_per_arm, 1, p_treat)

  # The analysis: an uncorrected two-sample test of proportions. `correct =
  # FALSE` turns OFF the continuity correction, which matters for two reasons.
  # It is what power.prop.test() assumes, and it is what power_exact() below
  # enumerates — so all three refer to the same test and can be compared.
  # suppressWarnings() silences the chi-square approximation warning that
  # fires on sparse tables; those cases are handled explicitly on the next line.
  pv <- suppressWarnings(
    prop.test(c(sum(y1), sum(y0)), c(n_per_arm, n_per_arm), correct = FALSE)$p.value
  )

  # A table with no events at all (or no non-events) in either arm gives a 0/0
  # test statistic and a NaN p-value. There is genuinely no evidence against
  # the null in that situation, so treat it as a non-rejection. Letting NaN
  # through would silently drop the replicate from the average instead, which
  # would bias the power upward by discarding only uninformative trials.
  if (!is.finite(pv)) pv <- 1

  c(reject = as.numeric(pv < alpha_test))
}

## ---------------------------------------------------------------------------
## run_sim() — repeat a simulated trial many times and summarise
##
##   n_sim    number of hypothetical trials to run
##   seed     integer fixing the random number streams
##   sim_one  a zero-argument function returning one trial's result, i.e. a
##            closure wrapping simulate_one() with its parameters filled in
##
## Returns a list with `reject` (the power), its Monte Carlo standard error,
## a 95% interval for it, the number of failed replicates, and the mean of
## everything else sim_one() returned.
## ---------------------------------------------------------------------------
run_sim <- function(n_sim, seed, sim_one) {

  # future_lapply() spreads the replicates across CPU cores. Passing an integer
  # to `future.seed` makes it build proper parallel random number streams from
  # that seed, so the answer is identical whether you run on 1 core or 10.
  # Using set.seed() with parallel workers would NOT give that guarantee.
  #
  # tryCatch returning NULL means one bad replicate cannot kill the whole run;
  # failures are counted below rather than crashing or being ignored.
  out <- future_lapply(seq_len(n_sim),
                       function(i) tryCatch(sim_one(), error = function(e) NULL),
                       future.seed = seed)

  # Separate the usable replicates from the failed ones. Reporting n_failed
  # matters: silently averaging over "whatever worked" would make a design
  # that only converges on easy scenarios look better than it is.
  ok  <- !vapply(out, is.null, logical(1))
  est <- colMeans(do.call(rbind, out[ok]))   # mean of every returned quantity

  # The power is the mean of `reject`, i.e. the proportion of simulated trials
  # that reached significance.
  power <- unname(est[["reject"]])
  ci <- wilson_ci(power, sum(ok))

  c(as.list(est), list(mc_se = sqrt(power * (1 - power) / sum(ok)),
                       lo = ci[1], hi = ci[2], n_failed = sum(!ok)))
}

## ---------------------------------------------------------------------------
## wilson_ci() — confidence interval for a simulated proportion
##
## This quantifies the SIMULATION's precision, not the trial's. It answers
## "if I re-ran these 20,000 replicates with a different seed, how much would
## the power estimate move?" — which is what tells you whether a difference
## between two scenarios is real or just noise.
##
## Wilson rather than the textbook Wald interval (p ± 1.96·SE) because Wald
## collapses to a single point when power hits 0 or 1. That is exactly where
## the validation below needs a sensible interval, since several scenarios sit
## at power ≈ 0.99.
## ---------------------------------------------------------------------------
wilson_ci <- function(p, n, z = 1.96) {
  d      <- 1 + z^2 / n
  centre <- (p + z^2 / (2 * n)) / d
  half   <- z / d * sqrt(p * (1 - p) / n + z^2 / (4 * n^2))
  c(max(0, centre - half), min(1, centre + half))
}

## ---------------------------------------------------------------------------
## power_exact() — the true power of the test, with no simulation at all
##
## A trial with n participants per arm can only produce a finite number of
## outcomes: x1 events in the treated arm (0 to n1) crossed with x0 in the
## control arm (0 to n0). This function walks over EVERY one of those tables,
## works out which would reject, and adds up their probabilities. The result
## is exact — no Monte Carlo error, no normal approximation.
##
## That makes it the right yardstick for checking the simulation. Comparing
## against power.prop.test() instead would be weaker, because it is itself a
## normal approximation and runs up to ~0.02 conservative at n = 150 per arm;
## agreement with it cannot distinguish "the simulation is correct" from "both
## approximations are wrong in the same direction".
##
## Cost is an (n1+1) x (n0+1) grid, so it is cheap here (1001 x 1001) but not
## something to call at n = 50,000.
## ---------------------------------------------------------------------------
power_exact <- function(n1, n0, p1, p0, alpha) {
  x1 <- 0:n1; x0 <- 0:n0                      # every possible event count

  # Pooled proportion under the null, for each (x1, x0) pair. outer() builds
  # the full grid of combinations in one shot rather than looping.
  pbar <- outer(x1, x0, "+") / (n1 + n0)
  se   <- sqrt(pbar * (1 - pbar) * (1 / n1 + 1 / n0))

  # The test statistic for each table. This is the same pooled-variance z test
  # that prop.test(correct = FALSE) computes in simulate_one().
  z <- outer(x1 / n1, x0 / n0, "-") / se

  # is.finite() drops the degenerate tables where se = 0, matching the
  # non-rejection convention used in simulate_one().
  rej <- is.finite(z) & abs(z) > qnorm(1 - alpha / 2)

  # Probability of each table under the true rates, summed over the rejecting
  # ones. The two arms are independent, so the joint probability is a product.
  sum(outer(dbinom(x1, n1, p1), dbinom(x0, n0, p0))[rej])
}

## ---------------------------------------------------------------------------
## req_n() — required sample size per arm, from the closed-form formula
##
## Inverts power.prop.test(): instead of "what power does n give?", asks "what
## n gives 80% power?". Uses the normal approximation rather than the exact
## enumeration above, deliberately — this is the number that goes in a protocol
## and that a reviewer will reproduce by hand. It is mildly conservative, so
## the required n errs slightly large, which is the safe direction.
##
## ceiling() because you cannot enrol a fraction of a participant.
## ---------------------------------------------------------------------------
req_n <- function(p0, p1, alpha, target = 0.80) {
  ceiling(power.prop.test(p1 = p0, p2 = p1, sig.level = alpha,
                          power = target, alternative = "two.sided")$n)
}

## ---------------------------------------------------------------------------
## pct() — format a proportion as a percentage for the tables
##
## Exists because sprintf("%.0f%%", 37.5) gives "38%", and these tables are
## read row by row by clinicians choosing the row that matches their beliefs.
## A 7.5-point effect silently displayed as a 7-point one is the kind of error
## nobody catches. formatC(format = "fg", digits = 3) keeps the half-points;
## trimws() strips the padding formatC adds to align a vector.
## ---------------------------------------------------------------------------
pct <- function(x) paste0(trimws(formatC(x * 100, format = "fg", digits = 3)), "%")
```

## Table 1a — required n per arm, active vs. control

```{r t1a}
T1 <- expand.grid(control = controls, diff = diffs)
T1 <- T1[order(T1$control, T1$diff), ]
T1$active   <- T1$control + T1$diff
T1$n_a025   <- mapply(req_n, T1$control, T1$active, alpha_2)
T1$n_a0167  <- mapply(req_n, T1$control, T1$active, alpha_3)
```

```{r t1a-table, echo = TRUE}
kable(data.frame(
  Control = pct(T1$control),
  Active  = pct(T1$active),
  `Difference (pp)` = T1$diff * 100,
  `n/arm, alpha = 0.025`  = T1$n_a025,
  `n/arm, alpha = 0.0167` = T1$n_a0167,
  check.names = FALSE
), caption = "Required n per arm for 80% power, active vs. control.")
```

## Table 1b — achieved power at the budget (1000/arm)

Simulated end to end, and checked against the exact power rather than against
another approximation.

```{r t1b, cache = TRUE}
T1b <- T1
T1b$sim   <- NA_real_; T1b$mc_se <- NA_real_
T1b$lo    <- NA_real_; T1b$hi    <- NA_real_
T1b$exact <- NA_real_

for (i in seq_len(nrow(T1b))) {
  r <- run_sim(n_sim, seed + i,
               function() simulate_one(T1b$control[i], T1b$active[i], n_arm, alpha_2))
  T1b$sim[i]   <- r$reject; T1b$mc_se[i] <- r$mc_se
  T1b$lo[i]    <- r$lo;     T1b$hi[i]    <- r$hi
  T1b$exact[i] <- power_exact(n_arm, n_arm, T1b$active[i], T1b$control[i], alpha_2)
}
T1b$within <- T1b$exact >= T1b$lo & T1b$exact <= T1b$hi
```

```{r t1b-table, echo = FALSE}
kable(data.frame(
  Control = pct(T1b$control),
  Active  = pct(T1b$active),
  `Diff (pp)` = T1b$diff * 100,
  `Power (simulated)` = sprintf("%.3f", T1b$sim),
  `MC SE` = sprintf("%.3f", T1b$mc_se),
  `Power (exact)` = sprintf("%.3f", T1b$exact),
  Agree = ifelse(T1b$within, "yes", "no"),
  check.names = FALSE
), caption = "Power at 1000/arm, alpha = 0.025. Simulation vs. exact enumeration.")
```

```{r t1b-check, echo = FALSE, results = "asis"}
cat(sprintf(
"At %s replicates per scenario the Monte Carlo SE is about %.4f, which makes
this a sharp test: the exact value falls inside the simulation's 95%% interval
in **%d of %d** scenarios, with a largest absolute discrepancy of %.4f.

A 95%% interval is expected to miss about %.1f times in %d by chance alone, so
%d outside is unremarkable — it would take several before the agreement looked
suspect. What matters is that nothing is off by a *practically* meaningful
amount, and nothing is: %.4f on powers running from 0.5 to 0.99 would not move
a single sample size decision.\n",
format(n_sim, big.mark = ","), mean(T1b$mc_se), sum(T1b$within), nrow(T1b),
max(abs(T1b$sim - T1b$exact)), 0.05 * nrow(T1b), nrow(T1b),
nrow(T1b) - sum(T1b$within), max(abs(T1b$sim - T1b$exact))))
```

## Table 2 — active 1 vs. active 2

The secondary contrast. Gaps are smaller here, since both arms are active.

```{r t2}
active1s <- c(0.30, 0.35, 0.40, 0.45, 0.50)
gaps     <- c(0.030, 0.050, 0.075, 0.100)

T2 <- expand.grid(active1 = active1s, gap = gaps)
T2 <- T2[order(T2$active1, T2$gap), ]
T2$active2   <- T2$active1 + T2$gap
T2$n_a05     <- mapply(req_n, T2$active1, T2$active2, alpha)
T2$n_a0167   <- mapply(req_n, T2$active1, T2$active2, alpha_3)
T2$pow_a05   <- mapply(function(a, b) power.prop.test(n_arm, a, b, alpha)$power,
                       T2$active1, T2$active2)
T2$pow_a0167 <- mapply(function(a, b) power.prop.test(n_arm, a, b, alpha_3)$power,
                       T2$active1, T2$active2)
```

```{r t2-table, echo = FALSE}
kable(data.frame(
  `Active 1` = pct(T2$active1),
  `Active 2` = pct(T2$active2),
  `Gap (pp)` = T2$gap * 100,
  `n/arm, alpha = 0.05`   = T2$n_a05,
  `n/arm, alpha = 0.0167` = T2$n_a0167,
  `Power at 1000/arm, alpha = 0.05`   = sprintf("%.3f", T2$pow_a05),
  `Power at 1000/arm, alpha = 0.0167` = sprintf("%.3f", T2$pow_a0167),
  check.names = FALSE
), caption = "Active 1 vs. Active 2: required n and power at the budget.")
```

## Table 3 — subgroup power

At the budget each arm contributes **333 rural** and **667 urban** participants.
Subgroup power is the same calculation at a smaller n — no new machinery.

```{r t3, cache = TRUE}
baselines <- c(0.25, 0.30, 0.35, 0.40, 0.45, 0.50)
effects   <- c(0.050, 0.075, 0.100, 0.125)

T3 <- expand.grid(control = baselines, diff = effects)
T3 <- T3[order(T3$control, T3$diff), ]
T3$active <- T3$control + T3$diff

T3$rural <- mapply(function(p0, p1) power_exact(n_rural, n_rural, p1, p0, alpha_2),
                   T3$control, T3$active)
T3$urban <- mapply(function(p0, p1) power_exact(n_urban, n_urban, p1, p0, alpha_2),
                   T3$control, T3$active)
```

Exact enumeration is cheap at these sample sizes, so the subgroup table uses it
directly rather than simulating.

```{r t3-table, echo = FALSE}
kable(data.frame(
  Control = pct(T3$control),
  Active  = pct(T3$active),
  `Diff (pp)` = T3$diff * 100,
  `Rural (333/arm)` = sprintf("%.3f", T3$rural),
  `Urban (667/arm)` = sprintf("%.3f", T3$urban),
  check.names = FALSE
), caption = "Within-subgroup power at the budget, alpha = 0.025.")
```

```{r t3-plot, echo = FALSE}
pd <- rbind(
  data.frame(T3[c("control", "diff")], power = T3$rural, subgroup = "Rural (333/arm)"),
  data.frame(T3[c("control", "diff")], power = T3$urban, subgroup = "Urban (667/arm)")
)
ggplot(pd, aes(diff * 100, power, colour = factor(control * 100), group = control)) +
  geom_hline(yintercept = 0.80, linetype = 2, colour = "grey45") +
  geom_line() + geom_point(size = 1.2) +
  facet_wrap(~subgroup) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = "Absolute difference vs. control (percentage points)",
       y = "Power", colour = "Control\nrate (%)",
       caption = "Dashed line = 80% power. alpha = 0.025.") +
  theme_minimal(base_size = 11)
```

## Minimum detectable effect

```{r mde}
mde <- function(n, p0, alpha, target = 0.80) {
  uniroot(function(d) power.prop.test(n, p0, p0 + d, alpha)$power - target,
          c(0.005, 0.40))$root
}
mde_tab <- data.frame(
  Control = pct(controls),
  `Full arm (1000)` = sprintf("%.1f", sapply(controls, function(p) mde(n_arm, p, alpha_2) * 100)),
  `Urban (667)`     = sprintf("%.1f", sapply(controls, function(p) mde(n_urban, p, alpha_2) * 100)),
  `Rural (333)`     = sprintf("%.1f", sapply(controls, function(p) mde(n_rural, p, alpha_2) * 100)),
  check.names = FALSE
)
```

```{r mde-table, echo = FALSE}
kable(mde_tab, row.names = FALSE,
      caption = "Minimum detectable difference (percentage points) at 80% power, alpha = 0.025.")
```

## Headline for the LOI

```{r headline, echo = FALSE, results = "asis"}
cat(sprintf(
"> With 3000 participants randomised 1:1:1, each active-vs-control comparison has
80%% power to detect an absolute difference of **%.1f percentage points** against
a 40%% control event rate (two-sided alpha = 0.025, Bonferroni-adjusted for two
primary comparisons). Within the prespecified subgroups, the corresponding
minimum detectable differences are **%.1f pp** in the urban stratum (667/arm) and
**%.1f pp** in the rural stratum (333/arm).\n",
mde(n_arm, 0.40, alpha_2) * 100,
mde(n_urban, 0.40, alpha_2) * 100,
mde(n_rural, 0.40, alpha_2) * 100))
```

## Assumptions and limitations

- Control event rates are a range rather than a point estimate, which is why
  every table varies them — the LOI does not commit to one.
- No allowance for dropout. If the outcome can be missing, inflate N
  accordingly or add it to the simulation.
- The simulated test is the uncorrected two-sample comparison of proportions.
  If the SAP specifies a covariate-adjusted model, power will differ somewhat.
- Nothing here represents site effects, clustering by clinician, or interim
  analyses.

## Notes for reuse

- **Unequal allocation.** Give `simulate_one()` separate arm sizes and pass both
  to `power_exact()`, which already takes `n1` and `n0`.
- **A different test.** Swap `prop.test()` for `fisher.test()` or a logistic
  model — but note the exact enumeration is specific to the uncorrected
  chi-square test, so it stops being the right yardstick.
- **Subgroups** need nothing new: run the same calculation at the stratum size.

## Reproducibility

```{r session, echo = FALSE}
sessionInfo()$R.version$version.string
```
