---
title: "PCORI LOI — three-arm pragmatic trial, binary outcome"
subtitle: "Sample size, achieved power at budget, and subgroup power"
author: "Zhe Chen"
date: "September 2026"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: kable
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 4.5)
source("../../R/load.R")
library(knitr)
```

## The question

Three arms — one control, two *different* active interventions — with a fixed
budget of **N = 3000**, so 1000 per arm. The clinicians need to read two things
straight off a table:

1. How large does an arm have to be to detect a given improvement over control?
2. At the budget we actually have, what effect can we detect — overall, and
   within the rural and urban subgroups?

The two active arms are **not** assumed to have the same effect. Every table
below varies one arm at a time, so a reader can pick the row matching whatever
they believe about each intervention separately.

### Multiplicity

Two primary comparisons (each active vs. control) → Bonferroni **α = 0.025**.
Three pairwise comparisons, if the active-vs-active contrast is also treated as
confirmatory → **α = 0.0167**. Both are shown throughout; which one applies is a
protocol decision, not a statistical one.

```{r design}
d <- design_binary_parallel()
d
n_arm <- 1000   # 3000 total / 3 arms

# Percentages, without rounding 37.5% to "38%" -- these tables are read
# directly by clinicians choosing a row, so a half-point matters.
pct <- function(x) paste0(trimws(formatC(x * 100, format = "fg", digits = 3)), "%")
```

## Table 1a — required n per arm, active vs. control

```{r t1a}
controls <- seq(0.30, 0.50, by = 0.05)
diffs    <- c(0.050, 0.075, 0.100)

req_n <- function(p0, p1, alpha, target = 0.80) {
  ceiling(power.prop.test(p1 = p0, p2 = p1, sig.level = alpha,
                          power = target, alternative = "two.sided")$n)
}

T1 <- expand.grid(control = controls, diff = diffs)
T1 <- T1[order(T1$control, T1$diff), ]
T1$active <- T1$control + T1$diff
T1$n_a025  <- mapply(req_n, T1$control, T1$active, 0.025)
T1$n_a0167 <- mapply(req_n, T1$control, T1$active, 0.05 / 3)

kable(data.frame(
  Control = pct(T1$control),
  Active  = pct(T1$active),
  `Difference (pp)` = T1$diff * 100,
  `n/arm, alpha = 0.025`  = T1$n_a025,
  `n/arm, alpha = 0.0167` = T1$n_a0167,
  check.names = FALSE
), caption = "Required n per arm for 80% power, active vs. control.")
```

These use `power.prop.test`, i.e. the normal approximation. That is deliberate:
it is the convention in protocols and the number a reviewer will get if they
check by hand. It is also mildly conservative — the exact power of the test is
a little higher at these sample sizes — so the required *n* here errs on the
safe side, which is the right direction for a sample size table.

Table 1b below simulates the same design at the budget and compares it against
the exact power, which is where the DGP actually gets validated.

## Table 1b — achieved power at the budget (1000/arm)

Run through the library so the numbers are simulated end-to-end rather than
read off a formula. The `power_analytic` column is the **exact** power of the
test being simulated (enumerated over every possible 2×2 table), which is what
makes the agreement meaningful.

```{r t1b, cache = TRUE}
g1 <- scenarios(p = lapply(seq_len(nrow(T1)), function(i) c(T1$control[i], T1$active[i])))
g1$n_per_arm    <- n_arm
g1$alpha        <- 0.05
g1$adjust       <- "bonferroni"
g1$n_comparisons <- 2

r1 <- check_against_analytic(d, g1, n_sim = 20000, seed = 2026, verbose = FALSE)
```

```{r t1b-table, echo = FALSE}
kable(data.frame(
  Control = pct(T1$control),
  Active  = pct(T1$active),
  `Diff (pp)` = T1$diff * 100,
  `Power (simulated)` = sprintf("%.3f", r1$power),
  `MC SE` = sprintf("%.3f", r1$mc_se),
  `Power (exact)` = sprintf("%.3f", r1$power_analytic),
  Agree = ifelse(r1$within_mc, "yes", "no"),
  check.names = FALSE
), caption = "Power at 1000/arm, alpha = 0.025. Simulation vs. exact enumeration.")
```

```{r t1b-check, echo = FALSE, results = "asis"}
cat(sprintf(
"At 20,000 replicates per scenario the Monte Carlo SE is about %.4f, which makes this a sharp test: the exact value falls inside the simulation's 95%% interval in **%d of %d** scenarios, and the largest absolute discrepancy anywhere in the table is %.4f.

A 95%% interval is expected to miss about %.1f times in %d by chance alone, so %d outside is unremarkable; it would take several before the agreement looked suspect. The thing to check is that no scenario is off by a *practically* meaningful amount, and none is: %.4f on powers running from 0.5 to 0.99 would not move a single sample size decision. That is the evidence that the DGP, the test, and the Bonferroni adjustment are all doing what the table says they are.\n",
mean(r1$mc_se), sum(r1$within_mc), nrow(r1), max(r1$abs_diff),
0.05 * nrow(r1), nrow(r1), nrow(r1) - sum(r1$within_mc),
max(r1$abs_diff)))
```

## Table 2 — active 1 vs. active 2

The secondary/exploratory contrast. Gaps are smaller here, since both arms are
active.

```{r t2}
active1s <- c(0.30, 0.35, 0.40, 0.45, 0.50)
gaps     <- c(0.030, 0.050, 0.075, 0.100)

T2 <- expand.grid(active1 = active1s, gap = gaps)
T2 <- T2[order(T2$active1, T2$gap), ]
T2$active2 <- T2$active1 + T2$gap
T2$n_a05   <- mapply(req_n, T2$active1, T2$active2, 0.05)
T2$n_a0167 <- mapply(req_n, T2$active1, T2$active2, 0.05 / 3)
T2$pow_a05   <- mapply(function(a, b) power.prop.test(n_arm, a, b, 0.05)$power,
                       T2$active1, T2$active2)
T2$pow_a0167 <- mapply(function(a, b) power.prop.test(n_arm, a, b, 0.05/3)$power,
                       T2$active1, T2$active2)

kable(data.frame(
  `Active 1` = pct(T2$active1),
  `Active 2` = pct(T2$active2),
  `Gap (pp)` = T2$gap * 100,
  `n/arm, alpha = 0.05`   = T2$n_a05,
  `n/arm, alpha = 0.0167` = T2$n_a0167,
  `Power at 1000/arm, alpha = 0.05`   = sprintf("%.3f", T2$pow_a05),
  `Power at 1000/arm, alpha = 0.0167` = sprintf("%.3f", T2$pow_a0167),
  check.names = FALSE
), caption = "Active 1 vs. Active 2: required n and power at the budget.")
```

## Table 3 — subgroup power

The trial is 1/3 rural, 2/3 urban, so at the budget each arm contributes
**333 rural** and **667 urban** participants. Subgroup power is just the same
design at a smaller n — no new machinery needed, which is the point of keeping
`n_per_arm` a parameter.

```{r t3, cache = TRUE}
n_rural <- round(n_arm / 3)      # 333
n_urban <- round(n_arm * 2 / 3)  # 667

baselines <- c(0.25, 0.30, 0.35, 0.40, 0.45, 0.50)
effects   <- c(0.050, 0.075, 0.100, 0.125)

T3 <- expand.grid(control = baselines, diff = effects)
T3 <- T3[order(T3$control, T3$diff), ]
T3$active <- T3$control + T3$diff

g3 <- scenarios(p = lapply(seq_len(nrow(T3)), function(i) c(T3$control[i], T3$active[i])))
g3 <- do.call(rbind, lapply(c(n_rural, n_urban), function(nn) { g3$n_per_arm <- nn; g3 }))
g3$alpha <- 0.05; g3$adjust <- "bonferroni"; g3$n_comparisons <- 2

r3 <- run_power(d, g3, n_sim = 4000, seed = 3026, verbose = FALSE)
half <- nrow(T3)
```

```{r t3-table, echo = FALSE}
kable(data.frame(
  Control = pct(T3$control),
  Active  = pct(T3$active),
  `Diff (pp)` = T3$diff * 100,
  `Rural (333/arm)` = sprintf("%.3f", r3$power[1:half]),
  `Urban (667/arm)` = sprintf("%.3f", r3$power[(half + 1):(2 * half)]),
  check.names = FALSE
), caption = "Within-subgroup power at the budget, alpha = 0.025.")
```

```{r t3-plot, echo = FALSE}
library(ggplot2)
pd <- rbind(
  data.frame(T3, power = r3$power[1:half],                  subgroup = "Rural (333/arm)"),
  data.frame(T3, power = r3$power[(half + 1):(2 * half)],   subgroup = "Urban (667/arm)")
)
ggplot(pd, aes(diff * 100, power, colour = factor(control * 100), group = control)) +
  geom_hline(yintercept = 0.80, linetype = 2, colour = "grey40") +
  geom_line() + geom_point(size = 1.2) +
  facet_wrap(~subgroup) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = "Absolute difference vs. control (percentage points)",
       y = "Power", colour = "Control\nevent rate (%)",
       caption = "Dashed line = 80% power. alpha = 0.025.") +
  theme_minimal(base_size = 11)
```

## Minimum detectable effect

The single number the LOI actually needs.

```{r mde}
mde <- function(n, p0, alpha, target = 0.80) {
  uniroot(function(dd) power.prop.test(n, p0, p0 + dd, alpha)$power - target,
          c(0.005, 0.40))$root
}
grid <- expand.grid(control = controls,
                    Group = c("Full arm (1000)", "Rural (333)", "Urban (667)"))
grid$n <- c(n_arm, n_rural, n_urban)[as.integer(grid$Group)]
grid$mde_pp <- mapply(function(n, p0) mde(n, p0, 0.025) * 100, grid$n, grid$control)

kable(reshape(
  data.frame(Control = pct(grid$control),
             Group = as.character(grid$Group),
             MDE = sprintf("%.1f", grid$mde_pp)),
  idvar = "Control", timevar = "Group", direction = "wide"),
  row.names = FALSE,
  col.names = c("Control rate", "Full arm (1000)", "Rural (333)", "Urban (667)"),
  caption = "Minimum detectable difference (percentage points) at 80% power, alpha = 0.025.")
```

## Headline for the LOI

```{r headline, echo = FALSE, results = "asis"}
m_full  <- mde(n_arm,  0.40, 0.025) * 100
m_rural <- mde(n_rural, 0.40, 0.025) * 100
m_urban <- mde(n_urban, 0.40, 0.025) * 100
cat(sprintf(
"> With 3000 participants randomised 1:1:1, each active-vs-control comparison has 80%% power to detect an absolute difference of **%.1f percentage points** against a 40%% control event rate (two-sided alpha = 0.025, Bonferroni-adjusted for two primary comparisons). Within the prespecified subgroups, the corresponding minimum detectable differences are **%.1f pp** in the urban stratum (667/arm) and **%.1f pp** in the rural stratum (333/arm).\n",
m_full, m_urban, m_rural))
```

## Notes for reuse

- **Unequal allocation.** Pass `n_per_arm = c(n1, n2, n3)`; the exact power
  routine handles unequal arms directly.
- **If the outcome is clustered** (sites, clinicians), switch to
  `design_binary_cluster()` and supply an ICC — see the ICC column in the
  [index](../../README.md).
- **Non-inferiority or a different test.** `test = "fisher"` or
  `test = "logistic"` swap the analysis without touching the DGP.
- The exact-enumeration check (`check_against_analytic`) costs almost nothing
  at these sample sizes. Run it whenever the DGP changes.

```{r session, echo = FALSE}
sessionInfo()$R.version$version.string
```
