---
title: "PCORI LOI — three-arm pragmatic trial, binary outcome"
subtitle: "Sample size, achieved power at budget, and subgroup power"
author: "Zhe Chen"
date: "September 2026"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 4.5)
library(future.apply)
library(knitr)
library(ggplot2)
plan(multisession, workers = max(1, parallel::detectCores() - 1))
```

## The question

Three arms — one control and two different active interventions — with a budget of N = 3000, corresponding to 1000 participants per arm.

The main questions are:

1. How large does an arm have to be to detect a given improvement over control?
2. At the available budget, what treatment effects can be detected?
3. What is the corresponding power within the rural and urban subgroups?


### Multiplicity

Two primary comparisons, each active against control → Bonferroni-adjusted level
**α = 0.025**. If the active-versus-active contrast is also treated as
confirmatory, three pairwise comparisons → **α = 0.0167**. 

The multiplicity strategy is a protocol decision. The calculations below display both levels where useful.

## Assumptions

```{r params}
n_arm      <- 1000          # budget: 3000 total across 3 arms
alpha      <- 0.05
alpha_2    <- alpha / 2     # 0.025  — two primary comparisons
alpha_3    <- alpha / 3     # 0.0167 — three pairwise comparisons
n_sim      <- 20000
seed       <- 2026

# The trial is 1/3 rural, 2/3 urban, so each arm contributes:
n_rural <- round(n_arm / 3)       # 333
n_urban <- round(n_arm * 2 / 3)   # 667

# Event-rate scenarios
controls <- seq(0.30, 0.50, by = 0.05)
diffs    <- c(0.050, 0.075, 0.100)
```

## Helper functions

```{r helpers}
## -------------- Analytical sample size and power calculation ---------------
## req_n() — required sample size per arm from the closed-form formula for a two-sided two-sample proportion test.
##
## This uses the large-sample approximation implemented by power.prop.test(). 
## It is used for the main sample size tables.
## ---------------------------------------------------------------------------
req_n <- function(p0, p1, alpha, target = 0.80) {
  ceiling(power.prop.test(p1 = p0, p2 = p1, sig.level = alpha,
                          power = target, alternative = "two.sided")$n)
}

## ---------------------------------------------------------------------------
## pct() — format a proportion as a percentage for the tables
##
## Exists because sprintf("%.0f%%", 37.5) gives "38%", and these tables are
## read row by row by clinicians choosing the row that matches their beliefs.
## formatC(format = "fg", digits = 3) keeps the half-points;
## trimws() strips the padding formatC adds to align a vector.
## ---------------------------------------------------------------------------
pct <- function(x) paste0(trimws(formatC(x * 100, format = "fg", digits = 3)), "%")

## ------------------- Monte Carlo simulation --------------------------------
## simulate_one() — generate and analyse ONE hypothetical trial
##
##   p_control   true event rate in the control arm
##   p_treat     true event rate in the active arm (the alternative we power for)
##   n_per_arm   participants in each arm
##   alpha_test  the level this test uses, already Bonferroni-adjusted
##
## Returns a named vector that must contain `reject` — 1 if this simulated
## trial would have declared a significant difference, 0 if not. Averaging
## `reject` over many trials is the power. 
## ---------------------------------------------------------------------------
simulate_one <- function(p_control, p_treat, n_per_arm, alpha_test) {
  
  # The current data-generating model assumes: independent participants, binary outcome, 
  # no clustering, no covariate adjustment, no dropout or missing outcome. 
  # If the planned trial analysis changes, this is the line that should be modified.
  y0 <- rbinom(n_per_arm, size = 1, prob = p_control) 
  y1 <- rbinom(n_per_arm, size = 1, prob = p_treat)

  # The analysis: two-sided pooled two-sample test of proportions. `correct =
  # FALSE` turns OFF the continuity correction, which matters for two reasons.
  # It is what power.prop.test() assumes, and it is what power_exact() below
  # enumerates, so all three refer to the same test and can be compared.
  # suppressWarnings() silences the chi-square approximation warning that
  # fires on sparse tables; those cases are handled explicitly on the next line.
  pv <- suppressWarnings( 
  prop.test( 
    x = c(sum(y1), sum(y0)), 
    n = c(n_per_arm, n_per_arm), 
    correct = FALSE, 
    alternative = "two.sided" 
  )$p.value 
  )

  # If all observations across both arms have the same outcome, 
  # the pooled standard error is zero and the test statistic is undefined. 
  # There is genuinely no evidence against the null in that situation, so treat it as a non-rejection. 
  if (!is.finite(pv)) pv <- 1

  c(reject = as.numeric(pv < alpha_test))
}

## ---------------------------------------------------------------------------
## run_sim() — repeat a simulated trial many times and summarise
##
##   n_sim    number of hypothetical trials to run
##   seed     integer fixing the random number streams
##   sim_one  a zero-argument function returning one trial's result, i.e. a
##            closure wrapping simulate_one() with its parameters filled in
##
## Returns a list with `reject` (the power), its Monte Carlo standard error,
## a 95% interval for it, the number of failed replicates, and the mean of
## everything else sim_one() returned.
## ---------------------------------------------------------------------------
run_sim <- function(n_sim, seed, sim_one) {

  # future_lapply() spreads the replicates across CPU cores. Passing an integer
  # to `future.seed` makes it build proper parallel random number streams from
  # that seed, so the answer is identical whether you run on 1 core or 10.
  # Using set.seed() with parallel workers would not give that guarantee.
  #
  # tryCatch returning NULL means one bad replicate cannot kill the whole run;
  # failures are counted below rather than crashing or being ignored.
  out <- future_lapply(seq_len(n_sim),
                       function(i) tryCatch(sim_one(), error = function(e) NULL),
                       future.seed = seed)

  # Separate the usable replicates from the failed ones. Reporting n_failed
  # matters: silently averaging over "whatever worked" would make a design
  # that only converges on easy scenarios look better than it is.
  ok  <- !vapply(out, is.null, logical(1))
  est <- colMeans(do.call(rbind, out[ok]))   # mean of every returned quantity

  # The power is the mean of `reject`, i.e. the proportion of simulated trials
  # that reached significance.
  power <- unname(est[["reject"]])
  ci <- wilson_ci(power, sum(ok))

  c(as.list(est), list(mc_se = sqrt(power * (1 - power) / sum(ok)),
                       lo = ci[1], hi = ci[2], n_failed = sum(!ok)))
}

## ---------------------------------------------------------------------------
## wilson_ci() — confidence interval for a simulated proportion
##
## This quantifies the Monte Carlo uncertainty in the estimated power. It answers
## "if I re-ran these 20,000 replicates with a different seed, how much would
## the power estimate move?" — which is what tells you whether a difference
## between two scenarios is real or just noise.
##
## Wilson rather than the Wald interval (p ± 1.96·SE) because Wald
## collapses to a single point when power hits 0 or 1. That is exactly where
## the validation below needs a sensible interval, since several scenarios sit
## at power ≈ 0.99.
## ---------------------------------------------------------------------------
wilson_ci <- function(p, n, z = 1.96) {
  d      <- 1 + z^2 / n
  centre <- (p + z^2 / (2 * n)) / d
  half   <- z / d * sqrt(p * (1 - p) / n + z^2 / (4 * n^2))
  c(max(0, centre - half), min(1, centre + half))
}

## --------------------- Exact enumeration -----------------------------------
## power_exact() — exact enumeration of the rejection probability
## for the pooled two-sample proportion z-test used in simulate_one(). 
##
## A trial with n participants per arm can only produce a finite number of
## outcomes: x1 events in the treated arm (0 to n1) crossed with x0 in the
## control arm (0 to n0). This function walks over EVERY one of those tables,
## works out which would reject, and adds up their probabilities. 
## This is an exact finite-sample calculation for the specified rejection 
## rule. It has no Monte Carlo error and no normal approximation.
##
## Cost is an (n1+1) x (n0+1) grid.
## ---------------------------------------------------------------------------
power_exact <- function(n1, n0, p1, p0, alpha) {
  # every possible event count
  x1 <- 0:n1
  x0 <- 0:n0                      

  # Pooled proportion under the null, for each (x1, x0) pair. 
  pbar <- outer(x1, x0, "+") / (n1 + n0)      # outer() builds the full grid of combinations.
  se   <- sqrt(pbar * (1 - pbar) * (1 / n1 + 1 / n0))  # Pooled null standard error.

  # The test statistic for each table. This is the same pooled-variance z test
  # that prop.test(correct = FALSE) computes in simulate_one().
  z <- outer(x1 / n1, x0 / n0, "-") / se

  # is.finite() drops the degenerate tables where se = 0, matching the
  # non-rejection convention used in simulate_one().
  rej <- is.finite(z) & abs(z) > qnorm(1 - alpha / 2)

  # Probability of each table under the true rates, summed over the rejecting
  # ones. The two arms are independent, so the joint probability is a product.
  sum(outer(dbinom(x1, n1, p1), dbinom(x0, n0, p0))[rej])
}
```

## Table 1a — required n per arm, active vs. control

The analytical calculation is used to determine the sample size required for 80% power.

```{r t1a}
T1 <- expand.grid(control = controls, diff = diffs)
T1 <- T1[order(T1$control, T1$diff), ]
T1$active   <- T1$control + T1$diff
T1$n_a025   <- mapply(req_n, T1$control, T1$active, alpha_2)
T1$n_a0167  <- mapply(req_n, T1$control, T1$active, alpha_3)
```

```{r t1a-table, echo = TRUE}
kable(data.frame(
  Control = pct(T1$control),
  Active  = pct(T1$active),
  `Difference (pp)` = T1$diff * 100,
  `n/arm, alpha = 0.025`  = T1$n_a025,
  `n/arm, alpha = 0.0167` = T1$n_a0167,
  check.names = FALSE
), caption = "Required n per arm for 80% power, active vs. control.")
```

## Table 1b — achieved power at the available budget (1000/arm)

The analytical calculation provides the main power calculation. 
Monte Carlo simulation is used to verify that the implementation of the planned two-sample test behaves as expected.
Exact enumeration provides an independent finite-sample benchmark for the same test.

```{r t1b, cache = TRUE}
T1b <- T1
T1b$approx <- NA_real_
T1b$sim   <- NA_real_; T1b$mc_se <- NA_real_
T1b$lo    <- NA_real_; T1b$hi    <- NA_real_
T1b$exact <- NA_real_

for (i in seq_len(nrow(T1b))) {

  # Analytical approximation
  T1b$approx[i] <- power.prop.test(n = n_arm, p1 = T1b$control[i], p2 = T1b$active[i], sig.level = alpha_2, alternative = "two.sided" )$power

  # Monte Carlo simulation
  r <- run_sim(n_sim, seed + i,
               function() simulate_one(T1b$control[i], T1b$active[i], n_arm, alpha_2))
  T1b$sim[i]   <- r$reject
  T1b$mc_se[i] <- r$mc_se
  T1b$lo[i]    <- r$lo
  T1b$hi[i]    <- r$hi

  # Monte Carlo simulation
  T1b$exact[i] <- power_exact(n_arm, n_arm, T1b$active[i], T1b$control[i], alpha_2)
}
T1b$within_mc_ci <- T1b$exact >= T1b$lo & T1b$exact <= T1b$hi
```

```{r t1b-table, echo = TRUE}
kable(data.frame(
  Control = pct(T1b$control),
  Active  = pct(T1b$active),
  `Diff (pp)` = T1b$diff * 100,
  `Power (analytical)` = sprintf("%.3f", T1b$approx),
  `Power (simulated)` = sprintf("%.3f", T1b$sim),
  `MC SE` = sprintf("%.3f", T1b$mc_se),
  `Power (exact)` = sprintf("%.3f", T1b$exact),
  # Agree = ifelse(T1b$within_mc_ci, "yes", "no"),
  check.names = FALSE
), caption = paste0( "Power at 1000/arm, two-sided alpha = ", alpha_2, "." ))
```
Simulation validation

```{r t1b-check, echo = FALSE, results = "asis"}
cat(sprintf(
"Across %s simulation replicates per scenario, the mean Monte Carlo SE was %.4f.
The exact power was within the simulation's 95%% Monte Carlo interval
in %d of %d scenarios, with a largest absolute discrepancy of %.4f.",

format(n_sim, big.mark = ","), mean(T1b$mc_se), sum(T1b$within_mc_ci), nrow(T1b),
max(abs(T1b$sim - T1b$exact))))
```

## Table 2 — active 1 vs. active 2

The active-vs-active comparison is treated here as a secondary/exploratory comparison at the nominal two-sided level of 0.05, with an additional column showing the result if all three pairwise comparisons were instead subject to Bonferroni adjustment.

```{r t2}
active1s <- c(0.30, 0.35, 0.40, 0.45, 0.50)
gaps <- c(0.030, 0.050, 0.075, 0.100)

T2 <- expand.grid(active1 = active1s, gap = gaps)
T2 <- T2[order(T2$active1, T2$gap), ]
T2$active2   <- T2$active1 + T2$gap
T2$n_a05     <- mapply(req_n, T2$active1, T2$active2, alpha)
T2$n_a0167   <- mapply(req_n, T2$active1, T2$active2, alpha_3)
T2$pow_a05 <- mapply(
  function(a, b)
    power.prop.test(
      n = n_arm,
      p1 = a,
      p2 = b,
      sig.level = alpha,
      alternative = "two.sided"
    )$power,
  T2$active1,
  T2$active2
)
T2$pow_a0167 <- mapply(function(a, b) power.prop.test(n_arm, a, b, alpha_3)$power,
                       T2$active1, T2$active2)
```

```{r t2-table, echo = TRUE}
kable(data.frame(
  `Active 1` = pct(T2$active1),
  `Active 2` = pct(T2$active2),
  `Gap (pp)` = T2$gap * 100,
  `n/arm, alpha = 0.05`   = T2$n_a05,
  `n/arm, alpha = 0.0167` = T2$n_a0167,
  `Power at 1000/arm, alpha = 0.05`   = sprintf("%.3f", T2$pow_a05),
  `Power at 1000/arm, alpha = 0.0167` = sprintf("%.3f", T2$pow_a0167),
  check.names = FALSE
), caption = "Active 1 vs. Active 2: required n per arm (80% power) and analytical power at the budget.")
```

## Table 3 — subgroup power

At the budget each arm contributes approximately **333 rural** and **667 urban** participants.
Subgroup power is the same calculation at a smaller n.

```{r t3, cache = TRUE}
baselines <- c(0.25, 0.30, 0.35, 0.40, 0.45, 0.50)
effects   <- c(0.050, 0.075, 0.100, 0.125)

T3 <- expand.grid(control = baselines, diff = effects)
T3 <- T3[order(T3$control, T3$diff), ]
T3$active <- T3$control + T3$diff

T3$rural <- mapply(function(p0, p1) power.prop.test(n_rural, p0, p1, alpha_2)$power,
                   T3$control, T3$active)
T3$urban <- mapply(function(p0, p1) power.prop.test(n_urban, p0, p1, alpha_2)$power,
                   T3$control, T3$active)
```

```{r t3-table, echo = FALSE}
kable(data.frame(
  Control = pct(T3$control),
  Active  = pct(T3$active),
  `Diff (pp)` = T3$diff * 100,
  `Rural (333/arm)` = sprintf("%.3f", T3$rural),
  `Urban (667/arm)` = sprintf("%.3f", T3$urban),
  check.names = FALSE
), caption = "Analytical power within subgroups at the available budget, alpha = 0.025.")
```

```{r t3-plot, echo = TRUE}
pd <- rbind(
  data.frame(T3[c("control", "diff")], power = T3$rural, subgroup = "Rural (333/arm)"),
  data.frame(T3[c("control", "diff")], power = T3$urban, subgroup = "Urban (667/arm)")
)
ggplot(pd, aes(diff * 100, power, colour = factor(control * 100), group = control)) +
  geom_hline(yintercept = 0.80, linetype = 2, colour = "grey45") +
  geom_line() + geom_point(size = 1.2) +
  facet_wrap(~subgroup) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = "Absolute difference vs. control (percentage points)",
       y = "Power", colour = "Control\nrate (%)",
       caption = paste0( "Dashed line = 80% power. Two-sided alpha = ", alpha_2, "." )) +
  theme_minimal(base_size = 11)
```

## Minimum detectable effect

The minimum detectable difference is the smallest positive absolute difference that gives the specified target power under the analytical approximation.

```{r mde}
mde <- function(n, p0, alpha, target = 0.80) {
  uniroot(function(d) power.prop.test(n, p0, p0 + d, alpha)$power - target,
          c(0.0001, 1 - p0 - 0.0001))$root
}
mde_tab <- data.frame(
  Control = pct(controls),
  `Full arm (1000)` = sprintf("%.1f", sapply(controls, function(p) mde(n_arm, p, alpha_2) * 100)),
  `Urban (667)`     = sprintf("%.1f", sapply(controls, function(p) mde(n_urban, p, alpha_2) * 100)),
  `Rural (333)`     = sprintf("%.1f", sapply(controls, function(p) mde(n_rural, p, alpha_2) * 100)),
  check.names = FALSE
)
```

```{r mde-table, echo = FALSE}
kable(mde_tab, row.names = FALSE,
      caption = paste0( "Minimum detectable difference (percentage points) at 80% power, ", 
          "two-sided alpha = ", alpha_2, "." )
)
```

## Headline for the LOI

```{r headline, echo = FALSE, results = "asis"}
cat(sprintf(
"> With 3000 participants randomised 1:1:1, each active-vs-control comparison has
80%% power to detect an absolute difference of **%.1f percentage points** against
a 40%% control event rate (two-sided alpha = 0.025, Bonferroni-adjusted for two
primary comparisons). Within the prespecified subgroups, the corresponding
minimum detectable differences are **%.1f pp** in the urban stratum (667/arm) and
**%.1f pp** in the rural stratum (333/arm).\n",
mde(n_arm, 0.40, alpha_2) * 100,
mde(n_urban, 0.40, alpha_2) * 100,
mde(n_rural, 0.40, alpha_2) * 100))
```

## Notes for reuse

- **Unequal allocation.** Give `simulate_one()` separate arm sizes and pass both
  to `power_exact()`, which already takes `n1` and `n0`.
- **A different test.** Swap `prop.test()` for `fisher.test()` or a logistic
  model — but note the exact enumeration is specific to the uncorrected
  chi-square test, so it stops being the right yardstick.
- **Subgroups** need nothing new: run the same calculation at the stratum size.

